---
title: "Questions from 2024-08-01"
author: "Anya Ferguson, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r}
library(targets)

source("R/quarto_formatting.R")
```

## Type 1 A&E attendances

### Regional patterns in ICBs

Question: *You’ve discussed the 4-fold difference in % of attendances being MH specific across the 42 ICB areas – I just wondered if there were any regional patterns that you noticed in these ICBs? I am very conscious that this could just be because these areas have different rurality’s and possible different screening practices. But if we pulled these areas out with caveats would that be fair?*

From the map of attendance rates by ICB in Figure 12, there aren't strong regional patterns (for example a north-south divide or coastal-inland split). 

For the breakdowns by rural/urban we used the number of people in rural or urban LSOAs from [https://www.gov.uk/government/statistics/2011-rural-urban-classification-lookup-tables-for-all-geographies](https://www.gov.uk/government/statistics/2011-rural-urban-classification-lookup-tables-for-all-geographies) to calculate rates per 100,000 population (for example Figure 18). Further analysis on how to classify the rurality of an ICB wold be needed to answer this question, but as a very quick analysis: Aggregating the 2011 rural-urban LSOA data up to ICB level, calculating the percentage of the population in an urban area in an ICB and plotting it against the percentage of type 1 MH attendances does not suggest a relationship between an ICB's rurality and the % of attendances.

```{r}
# Each ICB has count of rural, count of urban - so can create a % of how urban 
# it is:
rural_perc <- tar_read(rural_by_icb) |>
  tidyr::pivot_wider(names_from = rural_urban, 
                     values_from = count, 
                     values_fill = 0) |>
  janitor::clean_names() |>
  dplyr::mutate(perc_urban = urban / (urban + rural))

# Can then correlate this with the type1_perc_mh_attends:
type1_perc_mh_attends <- tar_read(type1_perc_mh_attends)

type1_perc_mh_attends |>
  dplyr::filter(der_financial_year == "2023/24") |>
  dplyr::left_join(rural_perc, "icb_code") |>
  ggplot2::ggplot(ggplot2::aes(value, perc_urban)) +
  ggplot2::geom_point() +
  ggplot2::labs(x = "Percentage of MH attendances",
                y = "Percentage of population in urban areas",
                title = "ICBs in England, 2023/24") +
  ggplot2::theme_bw()
```

### Disorders percentages

Question: *There are some concerns about treatment of psychotic disorders particularly, would you be happy for us to present percentages for psychotic disorder, schizophrenia etc separately? Can I calculate the denominator from those listed in the table at the top of this section? Or are there some smaller disorders not listed in the table?*

The denominator cannot be calculated from Table 1. The full table of Type 1 attendances by SNOMED term for 2023/24 has 70 different SNOMED terms with the majority having counts of less than 5. Some of these terms are subcategories of the disorders showing in the top ten - for example "Chronic schizophrenia" or "Reactive depressive psychosis". Each SNOMED term would need to be manually classified/grouped together to get the true numbers of percentages for any disorder.

```{r}
#| eval: false
# Removing top ten slice from SNOMED table:
gt_causes <- tar_read(ae_diag) |> 
  dplyr::filter(der_financial_year == '2023/24', ec_department_type == 1) |> 
  dplyr::summarise(attends = sum(attends), .by = term) |>
  dplyr::rename(SNOMED_term = term) |> 
  dplyr::arrange(desc(attends)) #|> 
  #slice_head(n = 10) 

create_dt(gt_causes) 
```

## Walk-in and UTC attendances:

### UEC attendances known to MH services

Question: *Can I just check the rationale for this statement: Typically, around 50% of these patients presenting in urgent care are known to specialist mental health services, suggesting their mental health needs are less acute. Are you suggesting that the cohort is less likely to be those that have attempted to end their lives? Or are there other reasons for this. I’m just wondering whether people who haven’t accessed MH services are more likely to go to type 3 services as they may be perceived as less overwhelming environments and they’ve never been in crisis before.*

50% of patients attending urgent care for MH reasons are already known to specialist mental health services. We are unsure about the statement "suggesting their mental health needs are less acute" and will double check with Andy when he gets back from leave.

In our analysis, we did not look at why patients attend urgent care compared to Type 1, so can't say that this may be due to perceptions of less overwhelming environments or having been in crisis before. We suspect that patients are more likely to be referred or taken to AE other urgent care if they are known to MH services because of implications of severity, but have not looked into this.

### Socio-economic gradients

Question: *Can I just check whether this statement is correct as it refers to A&E not urgent treatment centres and is same as the section above: There is a very clear socio-economic gradient in MH presentation rates. IMD1 (most deprived) have rates over 3 times higher than IMD10 the least deprived. Similarly, patients from urban areas are nearly twice as likely to present in A&E with MH problems.*

The socio-economic gradient is similar for Type 1 and Type 3/4 settings. Figures 16, 17 and their table and Figures 29, 30 and their table show that rates for IMD1 is about 3 times IMD10 for Type 1 and Type 3/4 attendances respectively.

## Calls to NHS 111

### Attendances and calls

Question: *When you say rate of attendance in this section is this to A&E following a call with NHS 111?*

In our rush to complete the NHS 111 section we may have used the word attendances instead of calls. I have reworded the section for clarity. If that doesn't help, please let me know and I'll look more into it.