---
title: "Blog(s)"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

``` {r setup}
library(targets)
```

## Ideas (long-list)

State of Care:

-   On the data (linking EC, UC, 111); Classifying MH in ECDS/111;

-   On significant sub-groups

-   On ED waiting times

-   On the volume and cost trends (and variation)

-   On access differences (ambulance, arrival time)

-   On regular attenders

-   On diagnoses (and links to services?)

-   111 disposition to A&E

MHA detentions:

-   General intro/overview

-   On the data - 'crafting spells' and impact on LoS

-   Conversions from short to longer-term (and variation)

-   Detention revolving door...

-   Outcomes - pants data!

-   Distance inequality - males and females, age group trends

## Running order - SOC?

### 1.    State of Care overview

Each year the Care Quality Commission (CQC) produce a ['State of Care'](https://www.cqc.org.uk/publications/major-report/state-care) report for Health and Social Care in England. The report typically draws together CQC's assessment and survey data alongside other publicly and routinely available data on services and service users.

For their 2023/24 update report, one of the key areas of focus was how people in mental health crisis used services. However, there were some gaps in understanding of how emergency and urgent care services were being used by those with mental health difficulties, potentially shining a light on lack of provision in specialist and community services.

To support this analysis, we drew on data from the Emergency Care Dataset (ECDS) and NHS 111 data to explore the trends, patterns and variation in utilisation of these care access points. We linked patients to the Mental Health Services Dataset (MHSDS) to assess how many of the patients were known to specialist services already.

For Accident & Emergency (A&E) and Urgent Care Centre (UCC) attendances we used a basket of 1,500 SNOMED codes to identify mental health presentation (primary complaint) and for NHS 111 a sub-set of mental health-related symptoms for mental health-related calls. Dementia and substance-misuse conditions were explicitly excluded.

In 2023/24 there were 200k attendances to A&E with a chief complaint of mental health. There were 20k attendances to urgent care centres and 220k calls to NHS 111 services where mental health was the primary reason for contact. The direct cost of these contacts is estimated at >£36m, not including related ambulance call and conveyance costs or onward care.

Our analysis suggests that:

The likelihood of someone with a mental health problem contacting urgent and emergency care services varies greatly according to who they are and where they live. Females, those in adolescence, those in ‘other’ minority ethnic groups and those in deprived areas all access services at higher rates.

``` {r}
#| label: fig-type1-map
#| fig-cap: Attendance rate by ICB per 100,000 population, Type 1 A&E. All of England, 2023/24.

tar_read(cyp_icb_ed_map_2324)
```

Attendance rates to Type 1 units (emergency departments) are falling gradually over time but those to urgent treatment centres are increasing. Patients with mental health difficulties also access services differently – use ambulances more, arrive out of hours – and are also more likely to wait longer in A&E departments for assessment and treatment than those with physical health complaints.

While the majority of these patients are known to mental health providers, a substantial minority are not. Typically, the most common presenting conditions are depression and anxiety though material numbers do have more severe disorders.

Stay tuned for more detailed blogs on some of these findings....

### 2.    Utilisation - key sub-group variations (ED, UCC or a comparison)

PROMPT (co-pilot)
"write me a blog on difference in population sub groups using emergency services for mental health reasons"

**Understanding the Differences in Population Sub-Groups Using Emergency Services for Mental Health Reasons**
<br>
Mental health is a critical aspect of overall well-being, and access to emergency services plays a vital role in providing immediate care during crises. However, the utilization of these services varies significantly across different population sub-groups. This blog explores these differences, shedding light on the factors influencing emergency department (ED) visits for mental health reasons.

Age Differences
One of the most notable differences in the use of emergency services for mental health issues is age. Data shows that adults aged 18-44 have the highest rates of ED visits for mental health disorders, with a rate of 64.9 visits per 1,000 adults1. This rate decreases with age, dropping to 47.5 for adults aged 45-64 and further to 34.9 for those aged 65 and over2. Younger adults may face more acute stressors related to career, education, and social relationships, which could contribute to higher ED visit rates.

Gender Disparities
Gender also plays a significant role in the utilization of emergency services for mental health. Studies indicate that men and women experience and respond to mental health crises differently. For instance, men are more likely to visit EDs for substance-related disorders, while women are more likely to seek help for mood disorders and anxiety3. These differences can be attributed to various social, biological, and psychological factors.

Socioeconomic Status
Socioeconomic status (SES) is another critical factor influencing ED visits for mental health reasons. Individuals from lower SES backgrounds are more likely to use emergency services due to limited access to regular mental health care. Medicaid is often the primary payment source for these visits, highlighting the reliance on public health insurance among lower-income groups2. Financial stress and lack of access to preventive care can exacerbate mental health issues, leading to more frequent crises.

Racial and Ethnic Variations
Racial and ethnic disparities are evident in the use of emergency services for mental health. For example, during the COVID-19 pandemic, there were significant differences in ED visit rates among various racial and ethnic groups4. These disparities can be attributed to factors such as systemic racism, cultural stigma around mental health, and differences in access to healthcare services. Minority groups often face barriers to mental health care, resulting in higher reliance on emergency services during crises.

Impact of the COVID-19 Pandemic
The COVID-19 pandemic has further highlighted and exacerbated these disparities. The stress and isolation associated with the pandemic led to an increase in mental health-related ED visits across all population sub-groups4. However, the impact was more pronounced among certain groups, such as young adults and racial minorities, who already faced higher levels of stress and limited access to mental health resources.

Conclusion
Understanding the differences in how various population sub-groups use emergency services for mental health reasons is crucial for developing targeted interventions and policies. By addressing the unique needs of each group, we can improve access to mental health care and reduce the reliance on emergency services. This approach not only enhances individual well-being but also alleviates the burden on emergency departments, allowing them to provide better care for all patients.

### 3.    Access and pathways - which sub-groups behave differently and maybe why?

PROMPT (co-pilot)
"write me a blog on access challenges for patients with mental health challenges using emergency services"

**Access Challenges for Patients with Mental Health Using Emergency Services**
<br>
Accessing emergency services for mental health issues can be a daunting experience for many patients. Despite the critical role these services play in providing immediate care during crises, several barriers hinder effective access and treatment. This blog explores the key challenges faced by patients with mental health challenges when using emergency services.

Limited Availability of Mental Health Professionals
One of the most significant barriers is the shortage of mental health professionals in emergency departments (EDs). Many regions face a critical lack of psychiatrists and other mental health specialists, which can lead to long wait times and inadequate care1. This shortage is particularly acute in rural areas, where access to specialized care is often limited.

Stigma and Discrimination
Stigma surrounding mental health issues remains a pervasive problem. Patients often face discrimination and negative attitudes from healthcare providers, which can deter them from seeking help2. This stigma can result in patients feeling marginalized and receiving suboptimal care compared to those with physical health issues.

Inadequate Training of Emergency Staff
Emergency department staff may lack the necessary training to handle mental health crises effectively. This can lead to misdiagnosis, inappropriate treatment, and a lack of empathy towards patients with mental health challenges3. Comprehensive training programs are essential to equip ED staff with the skills needed to provide compassionate and effective care.

Overcrowded Emergency Departments
Overcrowding in EDs is a widespread issue that affects the quality of care for all patients, including those with mental health challenges. High patient volumes can lead to long wait times and rushed consultations, which are particularly detrimental for individuals in mental health crises3. This environment can exacerbate patients’ stress and anxiety, making it harder for them to receive the care they need.

Financial Barriers
Financial constraints are a significant barrier to accessing emergency mental health services. Many patients may lack adequate insurance coverage, leading to high out-of-pocket costs1. This financial burden can prevent individuals from seeking timely care, worsening their mental health conditions.

Lack of Continuity of Care
Emergency departments are often not equipped to provide ongoing mental health care. Once the immediate crisis is addressed, patients may be discharged without a clear follow-up plan or connection to long-term mental health services2. This lack of continuity can result in repeated ED visits and a cycle of crisis care without sustainable solutions.

Safety Concerns
Safety concerns in EDs can also pose challenges for patients with mental health issues. The chaotic and sometimes hostile environment of an emergency department can be overwhelming and triggering for individuals in crisis3. Ensuring a safe and supportive environment is crucial for effective mental health care.

Conclusion
Addressing these access challenges requires a multifaceted approach. Increasing the availability of mental health professionals, improving training for emergency staff, reducing stigma, and ensuring financial and logistical support are essential steps. By tackling these barriers, we can enhance the quality of care for patients with mental health challenges and ensure they receive the timely and compassionate support they need.

## Running order - MHA?

### 1.    Background and making the data work
### 2.    Conversion pathways

PROMPT (co-pilot)
"write me a blog on the conversion pathways under the mental health act in england"

**Conversion Pathways Under the Mental Health Act in England**
<br>
The Mental Health Act (MHA) in England provides a legal framework for the treatment and care of individuals with mental health disorders. One of the critical aspects of this framework is the conversion pathways, which outline the processes for transferring individuals between different types of care settings. This blog explores these pathways, focusing on their importance and the procedures involved.

Understanding Conversion Pathways
Conversion pathways under the MHA refer to the processes by which individuals with mental health needs are transferred from one type of care setting to another. These pathways ensure that patients receive the appropriate level of care and support, whether they are moving from a community setting to a hospital, from one hospital to another, or from a hospital back to the community.

Key Conversion Pathways
Transfer from Community to Hospital (Section 2 and Section 3)
Section 2: This section allows for the compulsory admission of an individual to a hospital for assessment for up to 28 days. It is used when there is a need for a thorough evaluation of the person’s mental health condition1.
Section 3: This section permits the compulsory admission for treatment for up to six months, which can be renewed. It is used when it is clear that the individual requires treatment that cannot be provided in the community1.
Transfer Between Hospitals (Section 19)
Section 19: This section covers the transfer of patients from one hospital to another. Transfers can occur for various reasons, including the need for specialized care or closer proximity to family2. The process involves coordination between the hospitals to ensure continuity of care.
Transfer from Hospital to Community (Section 17A - Community Treatment Orders)
Section 17A: This section allows for the use of Community Treatment Orders (CTOs), which enable patients to be treated in the community while still being subject to certain conditions. CTOs are designed to provide a balance between ensuring ongoing treatment and allowing patients to live more independently3.
Transfer of Prisoners to Hospital (Sections 47 and 48)
Section 47: This section allows for the transfer of sentenced prisoners to a hospital for treatment of a mental disorder. The transfer requires the approval of the Secretary of State for Justice4.
Section 48: This section applies to unsentenced prisoners and those on remand. It allows for their transfer to a hospital for treatment, also requiring the Secretary of State’s approval4.
Importance of Conversion Pathways
Conversion pathways are crucial for several reasons:

Continuity of Care: They ensure that patients receive consistent and appropriate care as they move between different settings.
Specialized Treatment: Pathways facilitate access to specialized treatment that may not be available in the initial care setting.
Legal and Ethical Compliance: They ensure that transfers are conducted in accordance with legal requirements and ethical standards, protecting patients’ rights.
Support for Recovery: By providing structured transitions, these pathways support patients’ recovery and reintegration into the community.
Challenges and Considerations
While conversion pathways are essential, they also present challenges:

Coordination: Effective communication and coordination between different care providers are crucial to avoid disruptions in care.
Patient Consent: Ensuring that patients are informed and involved in decisions about their care is vital for ethical practice.
Resource Availability: Adequate resources, including staffing and facilities, are necessary to support smooth transitions.
Conclusion
Conversion pathways under the Mental Health Act in England play a vital role in ensuring that individuals with mental health needs receive appropriate and continuous care. By understanding and effectively implementing these pathways, healthcare providers can enhance the quality of care and support the recovery and well-being of patients.

### 3.    Revolving door?

PROMPT (co-pilot)
"write me a blog on the revolving door of MHA detentions in England"

**The Revolving Door of Mental Health Act Detentions in England**
<br>
The term “revolving door” in mental health care refers to the repeated admissions and discharges of patients under the Mental Health Act (MHA) in England. This phenomenon highlights significant challenges within the mental health care system, where individuals often cycle through periods of crisis without receiving the sustained support they need. This blog delves into the causes, implications, and potential solutions to this issue.

Understanding the Revolving Door Phenomenon
The revolving door phenomenon involves patients being repeatedly detained under the MHA, often due to a lack of continuous and effective mental health care. These repeated detentions can be distressing for patients and place a significant burden on healthcare resources.

Causes of the Revolving Door
Insufficient Community Support One of the primary reasons for repeated detentions is the lack of adequate community-based mental health services. When patients are discharged from hospitals, they often return to environments that lack the necessary support systems to help them manage their conditions1. This gap in care can lead to relapse and subsequent readmission.
Shortage of Mental Health Professionals The shortage of mental health professionals, including psychiatrists, psychologists, and social workers, exacerbates the problem. This shortage means that patients may not receive the timely and consistent care they need to stay well in the community2.
Stigma and Discrimination Stigma surrounding mental health issues can prevent individuals from seeking help early, leading to crises that necessitate emergency detention. Additionally, discrimination within the healthcare system can result in suboptimal care, further contributing to the cycle of readmissions3.
Inadequate Aftercare Services Aftercare services, which are supposed to support patients post-discharge, are often underfunded and poorly coordinated. Section 117 of the MHA mandates that local authorities and clinical commissioning groups provide aftercare, but inconsistencies in service provision can leave patients without the necessary support2.
Implications of the Revolving Door
Patient Well-being Repeated detentions can be traumatic for patients, leading to feelings of helplessness and frustration. This cycle can undermine their trust in the healthcare system and hinder their recovery.
Healthcare System Strain The revolving door phenomenon places a significant strain on the healthcare system. Emergency departments and psychiatric units often operate at full capacity, and repeated admissions can divert resources from other patients in need1.
Economic Costs The financial burden of repeated detentions is substantial. Costs associated with hospital admissions, emergency services, and aftercare can add up, straining public health budgets.
Potential Solutions
Enhanced Community Services Investing in robust community mental health services is crucial. This includes increasing the availability of community mental health teams, crisis intervention services, and support groups to help patients manage their conditions outside of hospital settings.
Integrated Care Models Implementing integrated care models that combine physical and mental health services can provide more comprehensive support for patients. These models ensure that patients receive holistic care, addressing both their mental and physical health needs.
Improved Aftercare Coordination Ensuring that aftercare services are well-coordinated and adequately funded is essential. This includes clear communication between hospitals, community services, and patients to create effective care plans that support long-term recovery.
Reducing Stigma Public education campaigns and training for healthcare professionals can help reduce stigma and discrimination. By fostering a more understanding and supportive environment, individuals may be more likely to seek help early and avoid crises.
Conclusion
Addressing the revolving door of MHA detentions requires a multifaceted approach that includes enhancing community services, improving aftercare coordination, and reducing stigma. By tackling these issues, we can provide better support for individuals with mental health challenges and reduce the cycle of repeated detentions.

## Running order - DTT?

### 1.    Background and making the data work
### 2.    Differences between sub-groups

PROMPT (co-pilot)
"write me a blog on inequalities in access to mental health inpatient services. focus on distance from home."

**Inequalities in Access to Mental Health Inpatient Services: The Impact of Distance from Home**
<br>
Access to mental health inpatient services is a critical component of comprehensive mental health care. However, significant inequalities exist, particularly when it comes to the distance patients must travel to receive care. This blog explores how distance from home affects access to mental health inpatient services and the broader implications for patient outcomes and health equity.

The Challenge of Distance
For many individuals, the distance to the nearest mental health inpatient facility can be a significant barrier to accessing care. This issue is particularly pronounced in rural areas, where mental health services are often sparse and widely dispersed1. Patients living in these areas may face long travel times, which can delay treatment and exacerbate mental health conditions.

Impact on Patient Outcomes
Delayed Treatment Long distances can lead to delays in seeking and receiving treatment. When patients must travel far to access inpatient services, they may postpone seeking help until their condition becomes critical1. This delay can result in more severe symptoms and a longer recovery period.
Increased Stress and Anxiety The stress of traveling long distances can worsen mental health conditions. For patients already experiencing anxiety, depression, or other mental health issues, the added burden of travel can be overwhelming2. This stress can deter individuals from seeking necessary care.
Reduced Family Support Being far from home can limit the support patients receive from family and friends. Family involvement is crucial for recovery, providing emotional support and helping with treatment adherence3. When patients are admitted to distant facilities, maintaining this support network becomes challenging.
Socioeconomic Disparities
Distance-related barriers to mental health inpatient services often intersect with socioeconomic disparities. Individuals from lower-income backgrounds may lack the resources to travel long distances, such as reliable transportation or the financial means to cover travel costs1. This can lead to unequal access to care and poorer health outcomes for disadvantaged populations.

Potential Solutions
Telehealth Services Expanding telehealth services can help bridge the gap for those living far from inpatient facilities. Telehealth can provide initial assessments, follow-up care, and ongoing support, reducing the need for frequent travel4. While not a complete substitute for inpatient care, telehealth can complement traditional services and improve access.
Mobile Mental Health Units Mobile mental health units can bring services directly to underserved areas. These units can provide assessments, crisis intervention, and even some forms of treatment, reducing the need for patients to travel long distances2. By bringing care to the community, mobile units can help address geographic disparities.
Investment in Local Facilities Increasing investment in local mental health facilities can reduce the distance patients need to travel. Building more inpatient units in rural and underserved areas can ensure that individuals have access to care closer to home3. This investment can also create jobs and support local economies.
Community-Based Support Programs Strengthening community-based mental health programs can provide ongoing support for individuals after they are discharged from inpatient care. These programs can help patients manage their conditions locally, reducing the need for readmission and long-distance travel4.
Conclusion
Addressing the inequalities in access to mental health inpatient services requires a multifaceted approach. By reducing the distance patients must travel to receive care, we can improve treatment outcomes, reduce stress, and ensure that all individuals have equitable access to the mental health services they need. Investing in telehealth, mobile units, local facilities, and community-based programs are essential steps toward achieving this goal.

### 3.    Capacity pinch-point - falling admissions, rising distances?

PROMPT (co-pilot)
"why are mental health admissions falling in children and young people in england but more being admitted further from home"

**Capacity pinch-point**
<br>
The trend of falling mental health admissions among children and young people in England, coupled with an increase in those being admitted further from home, can be attributed to several factors:

Decrease in Admissions
Improved Community Services: There has been a concerted effort to enhance community-based mental health services, which can help manage conditions before they escalate to the point of requiring inpatient care1. These services include early intervention programs, school-based mental health support, and increased availability of outpatient services.
Preventive Measures: Increased focus on preventive measures and early intervention has likely contributed to the reduction in admissions. Programs aimed at identifying and addressing mental health issues early can prevent the need for inpatient care2.
Telehealth Services: The expansion of telehealth services has provided an alternative to inpatient care, allowing children and young people to receive support remotely. This can reduce the need for hospital admissions by providing timely and accessible care1.
Increase in Distance from Home
Bed Shortages: Despite the overall decrease in admissions, there is still a shortage of inpatient beds, particularly in certain regions. This shortage forces some patients to be admitted to facilities further from home3.
Specialized Care Needs: Some children and young people require specialized care that is not available locally. This necessitates admission to specialized units that may be located far from their homes4.
Regional Disparities: There are significant regional disparities in the availability of mental health services. Areas with fewer resources may have higher rates of out-of-area placements, as local facilities are unable to meet the demand5.
Policy and Funding Issues: Funding constraints and policy decisions can impact the availability of local inpatient services. In some cases, resources may be allocated in ways that do not adequately support local needs, leading to more distant placements6.
Implications
Family Support: Being admitted far from home can limit the support patients receive from family and friends, which is crucial for recovery. This can lead to feelings of isolation and hinder the healing process4.
Continuity of Care: Distance can disrupt the continuity of care, making it harder to coordinate between inpatient and community services. This can result in fragmented care and poorer outcomes5.
Financial and Logistical Burdens: Traveling long distances for care can impose significant financial and logistical burdens on families, adding stress to an already challenging situation6.
Conclusion
Addressing these issues requires a multifaceted approach, including increasing local inpatient capacity, enhancing community services, and ensuring equitable distribution of resources. By tackling these challenges, we can improve access to mental health care for children and young people, ensuring they receive the support they need closer to home.

## Anya's attempts

### Age and ED attendances blog

In our recent analysis for the Care Quality Commission (CQC), we explored the rates of attendance to A&E and UEC by the following age groups: 0-17 years, 18-21 years, 22-39 years, 40-64 years, 65-74 years and 75+ years.

```{r}
rate_attends_1821_2324 <- targets::tar_read(age_breakdowns)$type1_mh_attends |>
  dplyr::filter(der_financial_year == "2023/24",
                age_group == "18-21") |>
  dplyr::pull(value) |>
  janitor::round_half_up(2)
```

The adolescent/transitional age group, 18-21 years, consistently has the highest rates of presentation to A&E for mental health reasons (@fig-type1_attends-age). In 2023/24, `r rate_attends_1821_2324` per 100,000 18-21 year olds attended A&E for mental health reasons. Similar trends were found for UEC attendances, with 18-21 year olds having the highest rates of presentation across the years.

```{r}
#| label: fig-type1_attends-age
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by age group.
targets::tar_read(age_plot)$type1_mh_attends
```

In the CQC's 2024 [State of Care report](https://www.cqc.org.uk/publications/major-report/state-care), it's said that children and young people had a worse experience than adults for mental health care and that they faced long waiting times at all stages of care: initial assessments, diagnoses, medicines, and therapies. Statistics from the [Mental Health Foundation](https://www.mentalhealth.org.uk/explore-mental-health/statistics/people-seeking-help-diagnosed-mental-health-problems-statistics) show that 16-24 years olds were less likely to receive mental health treatment compared to other age groups.Being able to receive mental health care may lead to a worsening of their situations forcing people into crises and then requiring ED attendance.

There are potentially other reasons why this transitional age group may have higher rates of ED attendance though. This is traditionally when people may face new life circumstances, making new decisions and facing new stressors: leaving home, changes in education or employment, new financial independence, and so on.

The [Mental Health Foundation](https://www.mentalhealth.org.uk/explore-mental-health/statistics/stress-statistics) found that more 18-24 year olds report stress related to the pressure to succeed than other age groups. Almost half of 18-24 year olds said that they experience high stress from comparing themselves to others and only 7% said that they never felt overwhelmed or unable to cope in the past year.

Individuals in this transitional age group are also likely to be seeing their care transitioning between paediatric to adult MH teams. The [NCEPOD Inbetweeners report](https://www.ncepod.org.uk/2023transition/The%20Inbetweeners_summary%20report.pdf) found that there is no clear pathway between children's and adult's healthcare services for young people with chronic conditions. The [NHS Long-term plan](https://www.longtermplan.nhs.uk/) aims to better manage this transition in mental health care by creating a comprehensive mental health service for 0-25 year olds. On this the [Royal College of Psychiatrists](https://www.rcpsych.ac.uk/docs/default-source/improving-care/better-mh-policy/position-statements/ps03_22.pdf?sfvrsn=23c278e0_8) have outlined different ways this could be done and the challenges involved: Extending current CMHS to CYP up to 25 years old would require further investment, staffing and training, but a mixture of different models may help. Central to all of this is the development and evaluation of the transition with the young person and their carers - an overarching principle in [NICE guidelines](https://www.nice.org.uk/guidance/ng43) on managing the transition from children's to adult' health and social care services.



### Other notes

The 0-17 age group is the only one where current rates are higher than in 2019/20.

- transitional stage in life too - not in school, new stressor re rent, uni, work, money, (perceived loss too - wanting to be independent) loss of support groups (friends) or ...

  
  Over the past decade, the number of workers aged 16–34 years who reported that their mental health limits the type or amount of work they can do has more than quadrupled
  - quote from: https://www.mentalhealth.org.uk/explore-mental-health/statistics/mental-health-work-statistics
  - Based on: https://www.health.org.uk/publications/long-reads/what-we-know-about-the-uk-s-working-age-health-challenge

**difference in population sub groups using emergency services for mental health reasons - Anya's thoughts**

Socioeconomic variation

Across all years in our analysis, there was a very clear socio-economic gradient in mental health presentation rates to emergency services. Using the Index of Multiple Deprivation (IMD), people in the most deprived areas have rates over 3 times higher than the least deprived.

```{r}
targets::tar_read(ae_attends_imd)
```

- Clear socio-economic gradient
- 3 times higher
- Why?
  - Extra stressors for P in IMD1 cf IMD10
    - IMD has financial, environmental (link to green spaces being better) components, fewer opportunities, worse health outcomes
      - Would be interested to see rates by specific domains
      - Contribute to incidence of MH disorders AND excerbate so more likely to crisis
  - Number/Investment in MH services in the area
  
  
**Conversion Pathways Under the Mental Health Act in England - Anya's thoughts**

Under the Mental Health Act 1983, people with a mental disorder that puts themselves or others at risk can be detained, taken to hospital and treated without their agreement. There are different sections of the MHA that can be used with different rules on length of detention, but the focus of this blog is on the conversion of short-term sections outlined below:

- Section 4: This is an emergency admission and lasts up to 72 hours only.
- Section 5(2): Whilst already an in-patient in hospital for any reason, a patient may be detained by their doctor for up to 72 hours.
- Section 5(4): Whilst already an in-patient in hospital for a mental disorder, a patient may be detained by a registered nurse for up to 6 hours.
- Section 136: A detention by a police officer for up to 24 hours. To be detained under Section 136, the person must be in a public place and the detention starts when they are taken to a place of safety.

Detentions under these sections are short to allow for a MHA assessment to be undertaken by a medical practitioner. Based on this assessment, the patient may be detained under another section of the MHA or discharged **?can patients also be admitted/treated/referred with their consent - not a detention, but also not a discharge?**

In our analysis of MHA detentions between financial years 2019/20 and 2023/24, the majority of CYP detentions starting under Section 4 (66.88%) and Section 136 (74.49%) were not converted to another Section of the MHA. 

A low conversion rate of short-term to long-term sections suggests that individuals did not need to be detained further - that the person did not require hospitalisation or treatment against their will. This does not necessarily mean that the use of the short-term detention was inappropriate. The detention may have been necessary as a life-saving intervention - a way to remove the person from their immediate crisis, to move them to a place of safety and to provide reassurance. The subsequent assessment by the medical practitioner may have found that the person is already receiving the support and care they need generally or that they are in agreement about future treatment, and therefore the detention could be ended rather than converted to another section of the MHA.

**The Revolving Door of Mental Health Act Detentions in England - Anya's thoughts**

Within 12 months, almost half of all MHA detentions of CYP in England had a re-detention and almost a quarter had a mental health related hospital admission. 

The revolving door of MHA detentions suggests individuals often cycle through periods of crisis without receiving the sustained support they need. Repeated detentions can be distressing to patients and could point

The revolving door of MHA detentions also places a significant burden on healthcare resources. Our previous work on [inequities in CYP MH services](https://www.strategyunitwm.nhs.uk/publications/inequities-children-and-young-peoples-mental-health-services) and the [impact of the MHA reforms](https://www.strategyunitwm.nhs.uk/publications/estimating-impact-proposed-reforms-mental-health-act-workload-psychiatrists) highlight an increasin burden on mental health services. But if the number of re-detentions and re-admissions could be reduced, this burden could be lessened. 

In our analysis, we found that the percentage of CYP MHA re-detentions is greater for females compared to males, for 18-24 year olds compared to under 18s and for patients from areas with the lowest IMD decile compared to all other deciles. We also found that these groups had shorter lengths of stay for their detention spells. This could suggest that these groups are being discharged too early or without sufficient self or community care arrangements in place. 

The revolving door of MHA detentions is likely due to a lack of continuous and effective mental health care. The government's [plans to reform the Mental Health Act](https://www.gov.uk/government/news/better-care-for-mental-health-patients-under-major-reforms) suggests that "Discharge processes will also be reviewed more broadly and will include a safety management plan for the patient, to keep themselves and other safe". If patients are able to access mental health care, they may be less likely to reach crises points avoiding the need for a re-detention.

Mental health care needs to be prioritised to ensure all individuals receive the right support and treatments. In line with the government's shift from intervention to prevention, if routine mental health care is accessible to those that need it, we could prevent the individual being detained and re-detained.
