---
title: "Blog(s)"
format: html
editor: visual
---

## Ideas

State of Care:

-   On the data (linking EC, UC, 111); Classifying MH in ECDS/111;

-   On significant sub-groups

-   On ED waiting times

-   On the volume and cost trends (and variation)

-   On access differences (ambulance, arrival time)

-   On regular attenders

-   On diagnoses (and links to services?)

-   111 disposition to A&E

MHA detentions:

-   General intro/overview

-   On the data - 'crafting spells' and impact on LoS

-   Conversions from short to longer-term (and variation)

-   Detention revolving door...

-   Outcomes - pants data!

-   Distance inequality - males and females, age group trends

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
