---
title: "Mental Health attendances in urgent and emergency care settings"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)

source("R/quarto_formatting.R")
```

## Introduction

The Care Quality Commission (CQC) are looking to supplement their State of Care (SOC) report with relevant additional data and analysis. The latest state of care report can be found here <https://www.cqc.org.uk/publications/major-report/state-care/2022-2023>.

One of the key areas of interest is around perceived 'system failure' as measured by inappropriate or avoidable attendance to urgent and emergency care settings for people with mental health problems, particularly in children and young people.

This report investigates these cases specifically in people under the age of 25, and supplements the main report which covers all individuals.

## Data and methods

We summarised routine data from the Emergency Care Dataset (ECDS) and the Mental Health Services Dataset (MHSDS) via NCDR <https://www.ardengemcsu.nhs.uk/services/business-intelligence/ncdr/> according to specific population sub-groups and activity break-downs at Integrated Care Board (ICB) level, just for under 25s. The period covers the 5 financial years from 1st April 2019 to 31st March 2024 unless otherwise stated.

Patients were linked using common pseudonymised ID from ECDS to the MHSDS to establish whether they were known to mental health services (referred and/or treated) or not.

The presence of a presenting mental health problem in **A&E data** was determined by all descendant SNOMED codes (n=1580) for 'Mental Health Disorders' in either the Chief Complaint field *OR* the 1st Diagnosis field. Codes relating to dementia, memory & cognition OR substance misuse were excluded.

Denominator data on population numbers by sub-group (age, gender, imd decile, urban-rural) were sourced from ONS mid-year estimates at LSOA level grouped to 2023 ICB boundaries. As 2023 MYE has not yet been published for LSOA, the 2022 data has been rolled forward as a proxy. For ethnic category populations, the 2011 and 2021 census data has been used to impute linear estimates of the intervening and following years. In order to match 2001 categories used in SUS data <https://www.datadictionary.nhs.uk/data_elements/ethnic_category.html>, the Chinese category has been moved from 'Asian' to 'Other' in the 2011 and 2021 census population data <https://www.ons.gov.uk/census/census2021dictionary/variablesbytopic/ethnicgroupnationalidentitylanguageandreligionvariablescensus2021/ethnicgroup/classifications>.

In the detailed breakdowns sections, utilisation rates are presented by population sub-groups per 100,000 people under the age of 25 . These plots include a dashed black line for the average of the sub-groups (for example @fig-type1_attends-gender). A separate overview plot is provided to show the overall rates (for example @fig-type1-line).

Calendar year data has been fully attributed to the financial year in which it falls e.g. 2021 = 2021/22.

### Data quality notes for A&E attendances:

Chief complaint code (SNOMED): The current completion rate is 83% with 81% valid (correct/usable format) for 2024/25 which has improved from 78% complete and 75% valid in 2019/2020. The missing data in this field is more likely to represent more acute or complex conditions that are not self-reported or determinable prior to assessment.

Ethnic category codes: The current field completion is 98.2% overall with a validity rate (correct/usable format) of 91.5% for 2024/25; however there is significant variation in subgroups - for example range is 57% - 95% for some categories. Missing data is more likely to be from minority ethnic groups or where English is not first language therefore some of our rate calculations may understate the utilisation in these population groups.

## Summary

The likelihood of someone with a mental health problem contacting urgent and emergency care services varies greatly according to who they are and where they live. Females, those in adolescence, those in 'other' minority ethnic groups and those in deprived areas all access services at higher rates.

Attendance rates to Type 1 units (emergency departments) are falling gradually over time but those to urgent treatment centres are increasing. Patients with mental health difficulties also access services differently -- use ambulances more, arrive out of hours -- and are also more likely to wait longer in A&E departments for assessment and treatment than those with physical health complaints.

While the majority of these patients are known to mental health providers, a substantial minority are not. Typically, the most common presenting conditions are depression and anxiety though material numbers do have more severe disorders.

### Type 1 A&E attendances:

-   In 2023/24 there were around 5 million attendances to ED by children or young people, of which 64,000 had a primary diagnosis of mental health condition.
-   These MH attendances cost \> £9m (not including ambulatory costs).
-   Depression and anxiety disorder attendances account for around 70% of MH attendances in younger people. Dissociative convulsions attendances makers up slightly over 7%, and psychotic disorders make up over 6%. No other categories make up more than 5% of attendances each.
-   Nationally, more than 85% of children and young people presenting with MH attendances to A&E are known to mental health services. This also means that nearly 15% are not.
-   Among children and young people, substantially more patients with MH disorders are conveyed to A&E by ambulance than those with physical ailments. Among younger patients with MH disorders, during the period covered, the proportion at its lowest is just below 30%, whereas among younger patients with physical ailments it is never higher 16%.
-   Patients presenting with MH issues are much more likely to attend between the hours of 8pm and 6am than those with physical issues.
-   Fewer patients with MH complaints leave the department before their assessment or treatment is concluded than those with physical complaints. In 23/24, 5.7% of younger patients with physical complaints left before assessment or treatment, whereas for those with MH complaints the rate was only 3.6%
-   In 2023/24, those presenting with MH issues waited longer in the department for assessment, treatment and to discharge from A&E. The average ED 'pathway' was approximately 40 minutes longer for MH patients.
-   There were over 800 younger patients in England attending Emergency OR Urgent care departments 5 or more times during the latest 12 month period with a primary mental health complaint.
-   The rate of attendance in younger females is substantially higher than younger males, in every year since 2019/20. For example in 23/24, the rate in younger females was 515/100,000, whereas in younger males it was 263/100,000.
-   Consistently, the 18-24 year old group have a much rate of attendance compared to under 18s. The 18-24 year old group appears to be declining over time, although  even at its lowest in 23/24 was still 5 times the rate in the under 18s group.
-   Younger patients with MH issues from the White 'Other' ethnic categories (Chinese, Arab and other) present more frequently than other groups.
-   There is a very clear socio-economic gradient in MH presentation rates. IMD1 (most deprived) have attendance rates over 2 times higher than IMD10 (the least deprived).

### Walk-in and UTC attendances:

-   In 2023/24 there were around 2.4 million attendances to Type 3 or 4 departments, of which 6,000 had a primary diagnosis of mental health condition.
-   These MH attendances cost £0.55m (not including ambulatory costs).
-   Depression and anxiety disorder attendances account for nearly 85% of MH attendances in younger people. Panic disorder attendances makers up slightly over 6%, and psychotic disorders and eating disorder attendances make up slightly more than 2% of attendances each.
-   Typically, around 50% of these patients presenting in urgent care are known to mental health services.
-   The overwhelming majority (90%+) of both MH and non-MH younger patients convey themselves to the urgent care departments.
-   Younger patients presenting with MH issues are slightly more likely to attend between 7pm and 6am than those without MH issues.
-   Fewer younger patients with MH complaints leave the department before their assessment or treatment is concluded than those with physical complaints. For example, in 23/24, for younger patients without a MH complaint the rate was 3.6%, whereas for younger patients with a MH complaint, the figure was 2.1%
-   The rate of attendance by women is consistently substantially higher than men. For example, in 23/24 it was 44 per 100,000 amongst younger women, whereas it was 28 per 100,000 in younger men.
-   There are consistently  much higher rates of attendance in 18-24 year olds than under 18 year olds. For example among the years observed, the rate in 18-24 never dipped below 110 per 100,000, whereas in under 18 year olds the rate never exceed 20 per 100,000.
-   Patients with MH issues from 'Other' ethnic categories (Chinese, Arab and other) present more frequently than other groups.
-   Broadly, there is socio-economic gradient in MH presentation rates, with more deprived deciles having higher attendance rates than more privileged deciles, however during years most affected by the pandemic, the most deprived decile (IMD1) actually had substantially lower attendance rates than slightly less deprived deciles (such as IMD2, IMD3 and IMD4).

## Detailed analysis - Accident and Emergency attendance {#sec-type1-overview}

```{r ae_summary_bits, include = FALSE}

ae_all <- tar_read(cyp_ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all)/1000000

ae_all_cost <- tar_read(cyp_ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all_cost)/1000000000

ae_mh <- tar_read(cyp_ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh)/1000

ae_mh_cost <- tar_read(cyp_ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh_cost)/1000000

```

These attendances are those specified within the ECDS as 'Type 1' emergency department attendances and typically should be for those patients with life-threatening symptoms or injuries requiring immediate triage and treatment.

The average A&E attendance in England costs £280 with those conveyed by ambulance costing an additional £390 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

During 2023/24 there were `r round(ae_all,2)` million Type 1 attendances costing £ `r round(ae_all_cost,2)` billion. Of those attendances c`r round(ae_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(ae_mh_cost,2)` million.

```{r}
#| label: tbl-type1-snomed
#| tbl-cap: Top 10 SNOMED codes in A&E presentation. All of England, 2023/24.

gt_causes <- tar_read(cyp_ae_diag) |> 
  filter(der_financial_year == '2023/24', ec_department_type == 1) |> 
  summarise(attends = sum(attends), .by = term) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n = 10) 

create_dt(gt_causes) 
```

### Access and patient pathway {#sec-type1-pathway}

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport.

::: panel-tabset
##### All modes

```{r}
#| label: fig-type1-arrival
#| fig-cap: Arrival mode to Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_trans_barplot)
```

##### Trend

```{r}
#| label: fig-type1-arrival-trend
#| fig-cap: Arrival mode to Type 1 A&E. All of England, 2019/20 to 2023/24.
tar_read(cyp_ae_trans_trends)
```

##### Table

```{r}
#| label: tbl-type1-arrival-trend
#| tbl-cap: Arrival mode to Type 1 A&E. All of England, 2019/20 to 2023/24.
tar_read(cyp_ae_transp_trends) |>
  create_dt()
```
:::

#### Time of arrival

Patients presenting with mental health complaints are more likely to attend after 8pm and before 6am than their contemporaries with physical health complaints - this could be suggestive of lesser access to out-of-hours psychiatric and crisis support or the increased likelihood of crisis occurring when people are alone.

::: panel-tabset
##### Chart

```{r ae_arrive_time_plot}
#| label: fig-type1-toa
#| fig-cap: Hour of arrival to Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_toa_plot)
```

##### Table

```{r ae_arrive_time_table}
#| label: tbl-type1-toa
#| tbl-cap: Number and percentage of attendances by time of arrival and financial year.
tar_read(cyp_ae_toa_summary) |>
  create_dt()
```
:::

#### Leaving the department

SNOMED departure status codes allow us to identify patients that left the department before either assessment or treatment was completed.

::: panel-tabset
##### Chart

```{r ae_left_plot}
#| label: fig-type1-left
#| fig-cap: Left department before assessment or treatment complete. All of England, 2023/24.
tar_read(cyp_ed_left_chart)
```

##### Table

```{r ae_left_summary}
#| label: tbl-type1-leave
#| tbl-cap: Number of attendances and patients leaving the departments before assessment or treatment by financial year.
tar_read(cyp_ae_left_summary) |>
  create_dt()
```
:::

#### Waiting times in ED

The ECDS allows calculation of times from arrival to assessment, treatment, conclusion and to departure.

::: panel-tabset
##### Assessment

```{r ae_wait_assess_plot}
#| label: fig-type1-waiting-ass
#| fig-cap: Waiting time for assessment, Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_times_assess_plot)
```

##### Treatment

```{r ae_wait_treat_plot}
#| label: fig-type1-waiting-treat
#| fig-cap: Waiting time for treatment, Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_times_treat_plot)
```

##### Conclusion

```{r ae_wait_conclude_plot}
#| label: fig-type1-waiting-concl
#| fig-cap: Waiting time for conclusion, Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_times_conclude_plot)
```

##### Departure

```{r ae_wait_depart_plot}
#| label: fig-type1-waiting-depart
#| fig-cap: Waiting time to leave department, Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_times_depart_plot)
```

##### Table

```{r}
#| label: tbl-type1-waiting-times
#| tbl-cap: Waiting time for assessment, treatment, conclusion and to leave department, Type 1 A&E. All of England, 2023/24.
tar_read(cyp_ae_times_table) |>
  create_dt()
```
:::

#### Mental health frequent attendees

```{r freq_fly_figs, include = FALSE}
avg_all_attends <- tar_read(cyp_ae_freq) |> summarise(total = sum(all_pats)) |> pull(total)/5

avg_mh_attend <- tar_read(cyp_ae_freq) |> summarise(total = sum(mh_pats)) |> pull(total)/5

perc_mh <- round(avg_mh_attend/avg_all_attends*100,1)

avg_ff_attend <- tar_read(cyp_ae_freq) |> summarise(total = sum(mh_freqfly)) |> pull(total)/5

perc_freq <- round(avg_ff_attend/avg_mh_attend*100,1)

```

Frequent attendees can be defined as those attending 5 or more times in a 12 month period with a primary complaint/diagnosis of mental health disorders.

Over the past 5 years in England, on average, there are `r comma(round(avg_mh_attend))` patients presenting with a mental health problem. Of these around `r comma(round(avg_ff_attend))` ( `r perc_freq`% ) are those who attend frequently. This varies by ICB area.

::: panel-tabset
##### Overview

```{r ae_freq_icb_plot}
#| label: fig-type1-freq
#| fig-cap: Percentage of with 5+ attendances in year, Type 1 A&E. All of England, 2023/24.

tar_read(cyp_ae_freq_boxplot)

```

##### Summary

```{r}
#| label: tbl-type1-freq-summary
#| tbl-cap: Summary of percentage of with 5+ attendances over time.
tar_read(cyp_ae_freq_data) |>
  ungroup() |>
  rename(icb_code = icb23cd,
         value = perc_freq) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-type1-freq
#| tbl-cap: Percentage of Mental Health Type 1 patients attending 5+ times a year by ICB.
tar_read(cyp_ae_freq_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 2.5, 0.35))
```
:::

### Detailed breakdowns {#sec-type1-breakdowns}

#### Percentage of all Type 1 attendances with Mental Health as primary reason/diagnosis {#sec-type1-mh-attends}

@fig-type1-perc-mh suggests the median percentage of Type 1 attendances with Mental Health as the primary reason/diagnosis is comparable with pre-pandemic levels. Breakdowns by ICB are in @tbl-type1-perc and breakdowns by population demographics below.

::: panel-tabset
##### Overview

```{r}
#| label: fig-type1-perc-mh
#| fig-cap: Percentage of Mental Health Type 1 attendances over time.
tar_read(cyp_type1_mh_attends_boxplot)
```

##### Summary

```{r}
#| label: tbl-type1-perc-summary
#| tbl-cap: Summary of percentage of Mental Health Type 1 attendances over time.
tar_read(cyp_type1_perc_mh_attends) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-type1-perc
#| tbl-cap: Percentage of Mental Health Type 1 attendances by ICB.
tar_read(cyp_type1_mh_attends_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 2.5, 0.35))
```
:::

#### Percentage of MH Type 1 attendances known to MH services {#sec-type1-mh-known}

The percentages of Mental Health Type 1 attendances where the patient is already known to MH services varies over time (@fig-type1-mh-known-boxplot) and by ICB (@tbl-type1-mh-known).

::: panel-tabset
##### Overview

```{r}
#| label: fig-type1-mh-known-boxplot
#| fig-cap: Percentage of Mental Health Type 1 attendances known to MH services over time.
tar_read(cyp_type1_mh_known_boxplot)
```

##### Summary

```{r}
#| label: tbl-type1-mh-known-summary
#| tbl-cap: Summary of percentage of Mental Health Type 1 known to MH services over time.
tar_read(cyp_type1_perc_mh_known) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-type1-mh-known
#| tbl-cap: Percentage of Mental Health Type 1 attendances known to MH services by ICB.
tar_read(cyp_type1_mh_known_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(60, 90, 4)) 
```
:::

#### Utilisation rates, population sub-groups {#sec-type1-utilisation}

Breakdowns for the number of Type 1 attendances with MH as a primary reason/diagnosis are below.

##### Geographical variation

::: panel-tabset
##### Map

```{r icb_map_allmh}
#| label: fig-type1-map
#| fig-cap: Attendance rate by ICB per 100,000 population, Type 1 A&E. All of England, 2023/24.
tar_read(cyp_icb_ed_map_2324)
```

##### Table

```{r}
tar_read(cyp_icb_rates_ed) |> 
  filter(der_financial_year == '2023/24') |>
  dplyr::left_join(tar_read(icb_codes_names), "icb_code") |>
  get_standard_table_for_breakdowns()
```
:::

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-type1-line
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances. All of England.
tar_read(cyp_avg_type1_mh_attends_rate_plot)
```

###### Table

```{r }
tar_read(cyp_avg_type1_mh_attends_rate) |>
  get_standard_table_for_breakdowns()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r type1_attends_gender_plot}
#| label: fig-type1_attends-gender
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by gender.
tar_read(cyp_gender_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_gender_table}
tar_read(cyp_gender_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("gender")
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r type1_attends_age_plot}
#| label: fig-type1_attends-age
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by age group.
tar_read(cyp_age_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_age_table}
tar_read(cyp_age_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("age_group")
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r type1_attends_ethnic_plot}
#| label: fig-type1_attends-ethnic
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by ethnic category.
tar_read(cyp_ethnic_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_ethnic_table}
tar_read(cyp_ethnic_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("ethnic_category")
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart1

```{r type1_attends_imd_plot}
#| label: fig-type1_attends-imd
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by Index of Multiple Deprivation decile.
tar_read(cyp_imd_plot)$type1_mh_attends
```

###### Chart2

```{r type1_attends_imd_plot2}
#| label: fig-type1_attends-imd2
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by Index of Multiple Deprivation decile and year.
tar_read(cyp_ae_attends_imd)
```

###### Table

```{r type1_attends_imd_table}
tar_read(cyp_imd_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("imd_decile")
```
:::

##### Rural-urban variation

::: panel-tabset
###### Chart

```{r type1_attends_rural_plot}
#| label: fig-type1_attends-rural
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by rural and urban populations.
tar_read(cyp_rural_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_rural_table}
tar_read(cyp_rural_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("rural_urban")
```
:::
:::

## Detailed analysis - Walk-in and Urgent Care attendance {#sec-type34-overview}

```{r uec_summary_bits, include = FALSE}

uec_all <- tar_read(cyp_uec_summary) |> filter(der_financial_year == '2023/24') |> pull(all)/1000000

uec_all_cost <- tar_read(cyp_uec_summary) |> filter(der_financial_year == '2023/24') |> pull(all_cost)/1000000000

uec_mh <- tar_read(cyp_uec_summary) |> filter(der_financial_year == '2023/24') |> pull(mh)/1000

uec_mh_cost <- tar_read(cyp_uec_summary) |> filter(der_financial_year == '2023/24') |> pull(mh_cost)/1000000

```

These attendances are those specified within the ECDS as 'Type 3' urgent treatment centre and 'Type 4' walk-in centre attendances and should be typically self-conveyed patients requiring urgent assessment and treatment to prevent deterioration.

The average urgent care attendance in England costs £107 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

During 2023/24 there were `r round(uec_all,2)` million Type 3/4 attendances costing £ `r round(uec_all_cost,2)` billion. Of those attendances c`r round(uec_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(uec_mh_cost,2)` million.

```{r uec_top_ten_causes}
#| label: tbl-uec-snomed
#| tbl-cap: Top 10 SNOMED codes in UCC and WIC presentation. All of England, 2023/24.

gt_causes <- tar_read(cyp_ae_diag) |> 
  filter(der_financial_year == '2023/24', ec_department_type %in% c(3)) |> 
  summarise(attends = sum(attends), .by = term) |>
  mutate(perc = attends/sum(attends)) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n = 10) 

create_dt(gt_causes) 
```

### Access and patient pathway {#sec-type34-pathway}

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport.

::: panel-tabset
##### All modes

```{r uec_arrive_mode}
#| label: fig-uec_attends-mode
#| fig-cap: Percentage of attendances by arrival mode, Type 3/4 departments in England, 2023/24.
tar_read(cyp_uec_trans_barplot)
```

##### Trend

```{r uec_arrive_mode_trend}
#| label: fig-uec_attends-mode-trend
#| fig-cap: Percentage of attendances by arrival mode, Type 3/4 departments in England, 2019/20 to 2023/24.
tar_read(cyp_uec_trans_trends)
```

##### Table

```{r uec_arrive_mode_table}
#| label: tbl-uec-arrival
#| tbl-cap: Number and percentage of attendances by arrival mode and financial year.
tar_read(cyp_uec_transp_trends) |>
  create_dt()
```
:::

#### Time of arrival

::: panel-tabset
##### Chart

```{r uec_arrive_time_plot}
#| label: fig-uec-toa
#| fig-cap: Percentage of attendances by hour of arrival, all of England, 2019/20 to 2023/24.
tar_read(cyp_uec_toa_plot)
```

##### Table

```{r uec_arrive_time_table}
#| label: tbl-uec-toa
#| tbl-cap: Number and percentage of attendances by hour of arrival, 2019/20 to 2023/24.
tar_read(cyp_uec_toa_summary) |>
  create_dt()
```
:::

#### Leaving the department

SNOMED departure status codes allow us to identify patients that left the department before either assessment or treatment was completed.

::: panel-tabset
##### Chart

```{r uec_left_plot}
#| label: fig-uec_attends-left
#| fig-cap: Percentage of attendances left before assessment or treatment, Type 3/4 departments in England, 2023/24.
tar_read(cyp_uec_left_chart)
```

##### Table

```{r uec_left_summary}
#| label: tbl-uec-leave
#| tbl-cap: Number of attendances and patients leaving the departments before assessment or treatment by financial year.
tar_read(cyp_uec_left_summary) |>
  create_dt()
```
:::

### Detailed breakdowns {#sec-type34-breakdown}

#### Percentage of all UEC attendances with MH as primary reason/diagnosis {#sec-uec-mh-attends}

@fig-uec-perc-mh suggests greater percentage of Mental Health UEC attendances in 2023/24. This varies by ICB (@tbl-uec-perc) where some ICBs have consistently low or high percentages over time.

::: panel-tabset
##### Overview

```{r}
#| label: fig-uec-perc-mh
#| fig-cap: Percentage of Mental Health UEC attendances over time.
tar_read(cyp_uec_mh_attends_boxplot)
```

##### Summary

```{r}
#| label: tbl-uec-perc-summary
#| tbl-cap: Summary of percentage of Mental Health UEC attendances over time.
tar_read(cyp_uec_perc_mh_attends) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-uec-perc
#| tbl-cap: Percentage of Mental Health UEC attendances by ICB.
tar_read(cyp_uec_mh_attends_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 1.8, 0.25))
```
:::

#### Percentage of MH UEC attendances known to MH services {#sec-uec-mh-known}

@fig-uec-mh-known-boxplot suggests lower percentage of Mental Health UEC attends were known in 2023/24. This again varies by ICB (@tbl-uec-mh-known).

::: panel-tabset
##### Overview

```{r}
#| label: fig-uec-mh-known-boxplot
#| fig-cap: Percentage of Mental Health UEC attendances known to MH services over time.
tar_read(cyp_uec_mh_known_boxplot)
```

##### Summary

```{r}
#| label: tbl-type1-mh_known-summary
#| tbl-cap: Summary of percentage of Mental Health UEC attendances known to MH services over time.
tar_read(cyp_uec_perc_mh_known) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-uec-mh-known
#| tbl-cap: Percentage of Mental Health UEC attendances known to MH services by ICB.
tar_read(cyp_uec_mh_known_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(30, 90, 8)) 
```
:::

#### Utilisation rates, population sub-groups {#sec-type34-populations}

Breakdowns for the number of attendances to Urgent Care or Walk-in with MH as primary reason/diagnosis are below.

##### Geographical variation

::: panel-tabset
##### Map

```{r icb_map_allmh_uec}
#| label: fig-uec_attends-map
#| fig-cap: Attendances rate per 100,000 population, Type 3/4 departments by ICB, 2023/24.
tar_read(cyp_icb_uec_map_2324)
```

##### Table

```{r}
tar_read(cyp_icb_rates_uec) |> 
  filter(der_financial_year == '2023/24') |>
  dplyr::left_join(tar_read(icb_codes_names), "icb_code") |>
  get_standard_table_for_breakdowns()
```
:::

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-uec-line
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances. All of England.
tar_read(cyp_avg_uec_mh_attends_rate_plot)
```

###### Table

```{r }
tar_read(cyp_avg_uec_mh_attends_rate) |>
  get_standard_table_for_breakdowns()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r uec_attends_gender_plot}
#| label: fig-uec-attends-gender
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by gender.
tar_read(cyp_gender_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_gender_table}
tar_read(cyp_gender_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("gender")
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r uec_attends_age_plot}
#| label: fig-uec-attends-age
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by age group.
tar_read(cyp_age_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_age_table}
tar_read(cyp_age_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("age_group")
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r uec_attends_ethnic_plot}
#| label: fig-uec-attends-ethnic
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by ethnic category.
tar_read(cyp_ethnic_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_ethnic_table}
tar_read(cyp_ethnic_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("ethnic_category")
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart1

```{r}
#| label: fig-uec-attends-imd_1
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by Index of Multiple Deprivation decile.
tar_read(cyp_imd_plot)$uec_mh_attends
```

###### Chart2

```{r uec_attends_imd_plot2}
#| label: fig-uec-attends-imd_2
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by Index of Multiple Deprivation decile and year.
tar_read(cyp_uec_attends_imd)
```

###### Table

```{r uec_attends_imd_table}
tar_read(cyp_imd_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("imd_decile")
```
:::

##### Rural-urban variation

::: panel-tabset
###### Chart

```{r uec_attends_rural_plot}
#| label: fig-uec-attends-urban
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by rural or urban classification.
tar_read(cyp_rural_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_rural_table}
#| label: tbl-uec-attends-rural
#| tbl-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by rural and urban populations.
tar_read(cyp_rural_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("rural_urban")
```
:::
:::

##  {#sec-nhs111-overview}
