---
title: "Mental Health Act"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)

source("R/quarto_formatting.R")
```

## Introduction

## Data and methods

## Summary

## Detailed analysis

### Mental Health Act Re-detentions

The number of Mental Health Act (MHA) re-detentions within 12 months is defined as the number of subsequent MHA detention end-dates within 365 days of the first MHA detention end-date in a financial year.

For example, if a patient had the following MHA detention end dates:

  2024-04-05 \
  2024-08-07 \
  2025-04-02 \
  2025-04-16 \

Then the number of MHA re-detentions within 12 months for the financial years 2024/25 and 2025/26 would be 2 and 1, respectively. The percentage of re-detentions within 12 months for financial years 2024/25 and 2025/26 would be 100% and 50%, respectively.

```{r}
redet <- tar_read(cyp_redetentions) |>
  dplyr::rename(redetentions = attends)
```

About half of MHA detentions are re-detentions (@tbl-redetentions-number) but this varies by ICB (@fig-redetentions-mh).

```{r}
#| label: tbl-redetentions-number
#| tbl-cap: Number of MHA detentions and redetentions within 12 months.
tar_read(cyp_redetentions) |>
  dplyr::summarise(detentions = sum(detentions),
                   redetentions = sum(attends),
                   .by = der_financial_year) |>
  mutate(perc = janitor::round_half_up(redetentions / detentions, 4)) |>
  create_dt()
```

::: panel-tabset

##### Overview

```{r}
#| label: fig-redetentions-mh
#| fig-cap: Percentage of MHA redetentions within 12 months.
tar_read(cyp_redetentions_perc_boxplot) +
  ggplot2::labs(title = "Percentage of redetentions within 12 months",
                subtitle = "England 2019/20 to 2022/23 by ICB.")
```

##### Summary

```{r}
#| label: tbl-redetentions-summary
#| tbl-cap: Summary of percentage of MHA redetentions within 12 months.
tar_read(cyp_redetentions_perc) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-redetentions
#| tbl-cap: Percentage of MHA redetentions within 12 months by ICB.
tar_read(cyp_redetentions_perc_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 0.9, 0.125))
```

:::

#### Formal and Informal Mental Health Act Re-detentions

The majority or MHA re-detentions are formal with less than 1% being informal
or unknown.

```{r}
cyp_redetentions_legal_status_perc <- tar_read(cyp_redetentions_legal_status) |>
  dplyr::summarise(redetentions = sum(attends), .by = c(legal_status, 
                                                         der_financial_year)) |>
  dplyr::mutate(perc = janitor::round_half_up(redetentions / sum(redetentions), 4), 
                .by = der_financial_year) 

cyp_redetentions_legal_status_perc |>
  dplyr::select(-perc) |>
  tidyr::pivot_wider(names_from = legal_status, values_from = redetentions) |>
  create_dt()
```



#### Rates of Mental Health Act Re-detentions

::: panel-tabset

##### Overview

::: panel-tabset

###### Chart

```{r}
#| label: fig-redetentions-line
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months. All of England.
tar_read(cyp_avg_redetentions_rate_plot)
```

###### Table

```{r }
tar_read(cyp_avg_redetentions_rate) |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns()
```

:::

##### Gender variation

::: panel-tabset

###### Chart

```{r redetentions_gender_plot}
#| label: fig-redetentions-gender
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by gender.
tar_read(gender_plot)$cyp_redetentions
```

###### Table

```{r redetentions_gender_table}
tar_read(gender_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("gender")
```

:::

##### Age-group variation

::: panel-tabset

###### Chart

```{r redetentions_age_plot}
#| label: fig-redetentions-age
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by age group.
tar_read(cyp_age_plot)$cyp_redetentions
```

###### Table

```{r redetentions_age_table}
tar_read(cyp_age_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("age_group")
```

:::

##### Ethnic group variation

::: panel-tabset

###### Chart

```{r redetentions_ethnic_plot}
#| label: fig-redetentions-ethnic
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by ethnic category.
tar_read(ethnic_plot)$cyp_redetentions
```

###### Table

```{r redetentions_ethnic_table}
tar_read(ethnic_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("ethnic_category")
```

:::

##### Socioeconomic variation

::: panel-tabset

###### Chart1

```{r redetentions_imd_plot}
#| label: fig-redetentions-imd
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by Index of Multiple Deprivation decile.
tar_read(imd_plot)$cyp_redetentions
```

###### Chart2

```{r redetentions_imd_plot2}
#| label: fig-redetentions-imd2
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by Index of Multiple Deprivation decile and year.
tar_read(cyp_redetentions_imd)
```

###### Table

```{r redetentions_imd_table}
tar_read(imd_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("imd_decile")
```

:::


### Travel distance
