---
title: "Mental Health Act"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)

source("R/quarto_formatting.R")
```

## Introduction

## Data and methods

## Summary

## Detailed analysis

### Mental Health Act Re-detentions

The number of Mental Health Act (MHA) re-detentions within 12 months is defined as the number of subsequent MHA detention end-dates within 365 days of the first MHA detention end-date in a financial year.

For example, if a patient was detained under the MHA on the following dates:

2024-04-05\
2024-08-07\
2025-04-02\
2025-04-16\

Then the number of MHA re-detentions within 12 months for the financial years 2024/25 and 2025/26 would be 2 and 1, respectively. The percentage of re-detentions within 12 months for financial years 2024/25 and 2025/26 would be 100% and 50%, respectively.

```{r}
redet <- tar_read(cyp_redetentions) |>
  dplyr::rename(redetentions = attends)
```

About half of MHA detentions are re-detentions (@fig-redetentions-mh) but this varies by ICB (@tbl-redetentions).

::: panel-tabset
##### Overview

```{r}
#| label: fig-redetentions-mh
#| fig-cap: Percentage of redetentions within 12 months.
tar_read(cyp_redetentions_perc_boxplot) +
  ggplot2::labs(title = "Percentage of redetentions within 12 months",
                subtitle = "England 2019/20 to 2022/23 by ICB.")
```

##### Summary

```{r}
#| label: tbl-redetentions-summary
#| tbl-cap: Summary of percentage of redetentions within 12 months.
tar_read(cyp_redetentions_perc) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-redetentions
#| tbl-cap: Percentage of redetentions within 12 months by ICB.
tar_read(cyp_redetentions_perc_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 0.9, 0.125))
```
:::

#### Formal and Informal Mental Health Act Re-detentions

```{r}
legal_status_22_23 <- tar_read(cyp_redetentions_legal_status) |>
  dplyr::filter(der_financial_year == "2022/23") |>
  dplyr::mutate(perc = 100 * value)

formal_legal_status_22_23 <- legal_status_22_23 |>
  dplyr::filter(legal_status == "Formal") |>
  dplyr::pull(perc) |>
  janitor::round_half_up(2)

informal_legal_status_22_23 <- legal_status_22_23 |>
  dplyr::filter(legal_status == "Informal") |>
  dplyr::pull(perc) |>
  janitor::round_half_up(2)
```

The percentage of re-detentions within 12 months where the legal status of the first MHA detention was informal is consistently lower than when the legal status was formal or unknown (@fig-redetentions-status). In 2022/23, `r formal_legal_status_22_23`% of formal MHA detentions had re-detentions, compared to `r informal_legal_status_22_23`% of informal MHA detentions.

::: panel-tabset
##### Chart

```{r}
#| label: fig-redetentions-status
#| fig-cap: Percentage of redetentions within 12 months by legal status.
tar_read(cyp_redetentions_legal_status_plot)
```

##### Table

```{r}
#| label: tbl-redetentions-status
#| tbl-cap: Percentage of redetentions within 12 months by legal status.
tar_read(cyp_redetentions_legal_status) |>
  dplyr::rename(redetentions = attends,
                perc = value) |>
  create_dt()
```
:::

#### Rates of Mental Health Act Re-detentions

::: panel-tabset

##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-redetentions-line
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months. All of England.
tar_read(cyp_avg_redetentions_rate_plot)
```

###### Table

```{r }
tar_read(cyp_avg_redetentions_rate) |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r redetentions_gender_plot}
#| label: fig-redetentions-gender
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by gender.
tar_read(gender_plot)$cyp_redetentions
```

###### Table

```{r redetentions_gender_table}
tar_read(gender_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("gender")
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r redetentions_age_plot}
#| label: fig-redetentions-age
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by age group.
tar_read(cyp_age_plot)$cyp_redetentions
```

###### Table

```{r redetentions_age_table}
tar_read(cyp_age_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("age_group")
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r redetentions_ethnic_plot}
#| label: fig-redetentions-ethnic
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by ethnic category.
tar_read(ethnic_plot)$cyp_redetentions
```

###### Table

```{r redetentions_ethnic_table}
tar_read(ethnic_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("ethnic_category")
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart1

```{r redetentions_imd_plot}
#| label: fig-redetentions-imd
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by Index of Multiple Deprivation decile.
tar_read(imd_plot)$cyp_redetentions
```

###### Chart2

```{r redetentions_imd_plot2}
#| label: fig-redetentions-imd2
#| fig-cap: Rates per 100,000 population of the number of MHA redetentions within 12 months by Index of Multiple Deprivation decile and year.
tar_read(cyp_redetentions_imd)
```

###### Table

```{r redetentions_imd_table}
tar_read(imd_breakdowns)$cyp_redetentions |>
  dplyr::rename(redetentions = attends) |>
  get_standard_table_for_breakdowns("imd_decile")
```
:::

### Travel distance

#### Overview of metric

The Ward Location Distance from Home variable represents the geographical distance between a patient's residential address and the ward location where they received treatment. This metric is commonly used to assess the accessibility of healthcare services and inform population health planning. By understanding the distance between a patient's home and the healthcare facility, we can gain insights into the convenience and proximity of services, identify potential barriers to healthcare access, and inform resource allocation decisions. Additionally, this data can be used for research studies to examine the relationship between travel distance and patient outcomes.

**Definition:** This variable represents the geographical distance between a patient's residential address and the ward location where they received treatment. It is often used to assess the accessibility of healthcare services and inform population health planning.

**Purpose:**

-   Accessibility: Measures the convenience and proximity of healthcare facilities for patients.

-   Population Health: Informs resource allocation and planning by understanding the distribution of healthcare services in relation to population density.

-   Research: Can be used to study the impact of travel distance on patient outcomes and adherence to treatment.

**Interpretation:**

-   Lower values: Indicate greater accessibility and convenience for patients.

-   Higher values: Suggest potential barriers to healthcare access, such as long travel times or transportation difficulties.

**Potential Use Cases:**

-   Patient Experience: Analyzing the relationship between ward location distance and patient satisfaction or outcomes.

-   Health Disparities: Identifying disparities in healthcare access based on geographic location.

-   Service Planning: Informing the location of new healthcare facilities or the expansion of existing services.

#### **Calculation of 'Ward location distance from home'**

Derived from: SITE CODE (OF TREATMENT)\
\
NHS Digital guidance: Look up the postcode for this SiteCode in the ODS reference data file that holds the address for organisation sites. Then identify the Northings\\Eastings for this postcode by using the ONS NHS Postcode Directory file.

**Reference:** <https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/data-set-technical-guidance/technical-glossary>

#### Summary

#### Variation

##### ICB

##### Gender

##### Age-group

##### Ethnic group

##### Socioeconomic variation
