---
title: "Mental Health Act"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)

source("R/quarto_formatting.R")
```

## Introduction

## Data and methods

## Summary

## Detailed analysis

### Mental Health Act Detention Spell Length of Stay

A MHA detention spell can consist of consecutive MHA detentions so the length of stay is defined as the difference in time between the start of the first MHA detention in a spell to the spell discharge date.

In 2023/24, the distribution of MHA detention spells length of stay is right-skewed with the majority being 1 day long (@fig-los-histo). There is also a small peak in the distribution at about day 27, probably a result of detentions under section 2 of the MHA being limited to up to 28 days (@fig-los-histo-zoomed).

```{r}
llos_perc <- tar_read(cyp_llos_perc)
```

Of the `r llos_perc |> dplyr::pull(count)` MHA detention spells in 2023/24, `r llos_perc |> dplyr::pull(perc_llos) |> janitor::round_half_up(2)`% were longer than 365 days. Most of these long spells were due to consecutive detentions under Section 3 of the MHA.

::: panel-tabset
#### All LoS

```{r}
#| label: fig-los-histo
#| fig-cap: Histogram of lengths of MHA detention spells for 2023/24.
tar_read(cyp_los_histo)
```

#### LoS \< 30 days

```{r}
#| label: fig-los-histo-zoomed
#| fig-cap: Histogram of lengths of MHA detention spells less than 30 days for 2023/24.
tar_read(cyp_los_histo_zoomed)
```
:::

```{r}
los_median_23_24 <- tar_read(cyp_los_median) |>
  dplyr::filter(der_financial_year == "2023/24") |>
  dplyr::summarise(median = median(value)) |>
  dplyr::pull()
```

The median length of MHA detention spell has increased (@fig-los-mh) over time to `r los_median_23_24` days in 2023/24. This varies by ICB (@tbl-los-icb).

::: panel-tabset
##### Overview

```{r}
#| label: fig-los-mh
#| fig-cap: Median length of Mental Health Act detention spell.
tar_read(cyp_los_boxplot)  +
  ggplot2::labs(y = "Median length of MHA detention spell",
                title = "Median length of MHA detention spell",
                subtitle = "England 2019/20 to 2023/24 by ICB")
```

##### Summary

```{r}
#| label: tbl-los-summary
#| tbl-cap: Summary of median length of Mental Health Act detention spell.
tar_read(cyp_los_median) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-los-icb
#| tbl-cap: Median length of Mental Health Act detention spell by ICB.
tar_read(cyp_los_median_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 39, 5))
```
:::

Median length of stay for MHA detention spells are higher for males compared to females, higher for under 18s compared to 18-24 year olds and also show ethnic group and socioeconomic variation.

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-los-line
#| fig-cap: Median length of MHA detention over time.
tar_read(cyp_los_plot_overview)$plot
```

###### Table

```{r }
tar_read(cyp_los_plot_overview)$table |>
  create_dt()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r los_gender_plot}
#| label: fig-los-gender
#| fig-cap: Median length of MHA detention over time by gender.
tar_read(cyp_los_plot_gender)$plot
```

###### Table

```{r los_gender_table}
tar_read(cyp_los_plot_gender)$table |>
  create_dt()
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r los_age_plot}
#| label: fig-los-age
#| fig-cap: Median length of MHA detention over time by age group.
tar_read(cyp_los_plot_age_group)$plot
```

###### Table

```{r los_age_table}
tar_read(cyp_los_plot_age_group)$table |>
  create_dt()
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r los_ethnic_plot}
#| label: fig-los-ethnic
#| fig-cap: Median length of MHA detention over time by ethnic category.
tar_read(cyp_los_plot_ethnic_category)$plot
```

###### Table

```{r los_ethnic_table}
tar_read(cyp_los_plot_ethnic_category)$table |>
  create_dt()
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart

```{r los_imd_plot}
#| label: fig-los-imd
#| fig-cap: Median length of MHA detention over time by Index of Multiple Deprivation decile.
tar_read(cyp_los_plot_imd_decile)$plot
```

###### Table

```{r los_imd_table}
tar_read(cyp_los_plot_imd_decile)$table |>
  create_dt()
```
:::
:::

#### LOS by section of the Mental Health Act

As seen in @-ref-to-conversions-section-when-available, some spells start under one section of the MHA and are then converted to other sections. In this section of the report, the length of stay is explored by the section of the first MHA detention in the spell.

Extreme outliers have been excluded from some of the boxplots in this section for visualisation purposes. Charts have been annotated in this case and outliers are still present in the boxplot summary statistics.

##### Sections 135 or 136

```{r}
llos_perc_135_136 <- tar_read(cyp_llos_perc_135_136)
```

There were `r llos_perc_135_136 |> dplyr::pull(count)` spells that began with a MHA detention under sections 135 or 136 in 2023/24. Of these spells `r 100 - llos_perc_135_136 |> dplyr::pull(perc_llos) |> janitor::round_half_up(2)`%, were up to 1 day long.

```{r}
#| label: fig-los-histo_135_136
#| fig-cap: Histogram of lengths of MHA detention spells for 2023/24 where the first detention was under section 135 or 136.
tar_read(cyp_los_histo_135_136)
```

:::panel-tabset

##### Overview

```{r}
#| label: fig-los-mh_135_136
#| fig-cap: Median length of Mental Health Act detention spell where the first detention was under sections 135 or 136.
tar_read(cyp_los_boxplot_135_136)  +
  ggplot2::labs(y = "Median length of MHA detention spell",
                title = "Median length of MHA detention spell ",
                subtitle = "All spells that began with a sections 135 or 136 detention in England 2019/20 to 2023/24 by ICB",
                caption = "Extreme outliers have been excluded from this chart for visualisation purposes")
```

##### Summary

```{r}
#| label: tbl-los-summary_135_136
#| tbl-cap: Summary of median length of Mental Health Act detention spell where the first detention was under sections 135 or 136.
tar_read(cyp_los_median_135_136) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-los-icb_135_136
#| tbl-cap: Median length of Mental Health Act detention spell by ICB where the first detention was under sections 135 or 136.
tar_read(cyp_los_median_table_135_136) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 21, 3))
```
:::

##### Section 2

```{r}
llos_perc_2 <- tar_read(cyp_llos_perc_2)
```

There were `r llos_perc_2 |> dplyr::pull(count)` spells that began with a MHA detention under section 2 in 2023/24. Of these spells `r 100 - llos_perc_2 |> dplyr::pull(perc_llos) |> janitor::round_half_up(2)`%, were up to 28 days long.

```{r}
#| label: fig-los-histo_2
#| fig-cap: Histogram of lengths of MHA detention spells for 2023/24 where the first detention was under section 2.
tar_read(cyp_los_histo_2)
```

::: panel-tabset
##### Overview

```{r}
#| label: fig-los-mh_2
#| fig-cap: Median length of Mental Health Act detention spell where the first detention was under section 2.
tar_read(cyp_los_boxplot_2)  +
  ggplot2::labs(y = "Median length of MHA detention spell",
                title = "Median length of MHA detention spell ",
                subtitle = "All spells that began with a section 2 detention in England 2019/20 to 2023/24 by ICB",
                caption = "Extreme outliers have been excluded from this chart for visualisation purposes")
```

##### Summary

```{r}
#| label: tbl-los-summary_2
#| tbl-cap: Summary of median length of Mental Health Act detention spell where the first detention was under section 2.
tar_read(cyp_los_median_2) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-los-icb_2
#| tbl-cap: Median length of Mental Health Act detention spell by ICB where the first detention was under section 2.
tar_read(cyp_los_median_table_2) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 39, 5))
```
:::

##### Section 3

```{r}
llos_perc_3 <- tar_read(cyp_llos_perc_3)
```

There were `r llos_perc_3 |> dplyr::pull(count)` spells that began with a MHA detention under section 3 in 2023/24. Of these spells `r 100 - llos_perc_3 |> dplyr::pull(perc_llos) |> janitor::round_half_up(3)`%, were up to 365 days long.

```{r}
#| label: fig-los-histo_3
#| fig-cap: Histogram of lengths of MHA detention spells for 2023/24 where the first detention was under section 3.
tar_read(cyp_los_histo_3)
```

::: panel-tabset
##### Overview

```{r}
#| label: fig-los-mh_3
#| fig-cap: Median length of Mental Health Act detention spell where the first detention was under section 3.
tar_read(cyp_los_boxplot_3)  +
  ggplot2::labs(y = "Median length of MHA detention spell",
                title = "Median length of MHA detention spell ",
                subtitle = "All spells that began with a section 3 detention in England 2019/20 to 2023/24 by ICB")
```

##### Summary

```{r}
#| label: tbl-los-summary_3
#| tbl-cap: Summary of median length of Mental Health Act detention spell where the first detention was under section 3.
tar_read(cyp_los_median_3) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-los-icb_3
#| tbl-cap: Median length of Mental Health Act detention spell by ICB where the first detention was under section 3.
tar_read(cyp_los_median_table_3) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 365, 50))
```
:::

### Mental Health Act Re-detentions

The number of Mental Health Act (MHA) re-detentions within 12 months is defined as the number of subsequent MHA detention spell discharge dates within 365 days of the first MHA detention spell discharge date in a financial year.

For example, if a patient had the following MHA detention spell discharge dates:

2024-04-05\
2024-08-07\
2025-04-02\
2025-04-16\

Then the number of MHA re-detentions within 12 months for the financial years 2024/25 and 2025/26 would be 2 and 1, respectively. The percentage of MHA detentions that had a re-detention within 12 months for financial years 2024/25 and 2025/26 would be 100% and 50%, respectively.

```{r}
redet <- tar_read(cyp_redetentions) |>
  dplyr::rename(redetentions = attends)
```

About half of MHA detentions had a re-detention within 12 months (@tbl-redetentions-number) but this varies by ICB (@fig-redetentions-mh).

```{r}
#| label: tbl-redetentions-number
#| tbl-cap: Number of MHA detentions and redetentions within 12 months.
tar_read(cyp_redetentions_perc)$table |>
  create_dt()
```

::: panel-tabset
###### Overview

```{r}
#| label: fig-redetentions-mh
#| fig-cap: Percentage of MHA redetentions within 12 months.
tar_read(cyp_redetentions_perc_boxplot) +
  ggplot2::labs(title = "Percentage of redetentions within 12 months",
                subtitle = "England 2019/20 to 2022/23 by ICB.")
```

###### Summary

```{r}
#| label: tbl-redetentions-summary
#| tbl-cap: Summary of percentage of MHA redetentions within 12 months.
tar_read(cyp_redetentions_perc_icb_code) |>
  get_quantiles_and_mean()
```

###### ICB Variation

```{r}
#| label: tbl-redetentions
#| tbl-cap: Percentage of MHA redetentions within 12 months by ICB.
tar_read(cyp_redetentions_perc_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 90, 12.5))
```
:::

The percentage of MHA re-detentions is greater for females compared to males and higher for 18-24 years olds compared to under 18s. The percentage of MHA re-detentions is greater for patients from areas with the lowest IMD decile compared to all other IMD deciles.

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-redetention-line
#| fig-cap: Percentage of MHA redetentions within 12 months over time.
tar_read(cyp_redetentions_perc)$plot
```

###### Table

```{r }
tar_read(cyp_redetentions_perc)$table |>
  create_dt()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r redetentions_gender_plot}
#| label: fig-redetentions-gender
#| fig-cap: Percentage of MHA redetentions within 12 months by gender.
tar_read(cyp_redetentions_plot_gender)$plot
```

###### Table

```{r redetentions_gender_table}
tar_read(cyp_redetentions_plot_gender)$table |>
  create_dt()
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r redetentions_age_plot}
#| label: fig-redetentions-age
#| fig-cap: Percentage of MHA redetentions within 12 months by age group.
tar_read(cyp_redetentions_plot_age_group)$plot
```

###### Table

```{r redetentions_age_table}
tar_read(cyp_redetentions_plot_age_group)$table |>
  create_dt()
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r redetentions_ethnic_plot}
#| label: fig-redetentions-ethnic
#| fig-cap: Percentage of MHA redetentions within 12 months by ethnic category.
tar_read(cyp_redetentions_plot_ethnic_category)$plot
```

###### Table

```{r redetentions_ethnic_table}
tar_read(cyp_redetentions_plot_ethnic_category)$table |>
  create_dt()
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart

```{r redetentions_imd_plot}
#| label: fig-redetentions-imd
#| fig-cap: Percentage of MHA redetentions within 12 months by Index of Multiple Deprivation decile.
tar_read(cyp_redetentions_plot_imd_decile)$plot
```

###### Table

```{r redetentions_imd_table}
tar_read(cyp_redetentions_plot_imd_decile)$table |>
  create_dt()
```
:::
:::

#### Formal and Informal Mental Health Act Re-detentions

The majority or MHA re-detentions are formal with less than 1% being informal or unknown.

```{r}
tar_read(cyp_redetentions_formal_table) |>
  create_dt()
```

#### Re-admissions after a Mental Health Act detention

A re-admission is defined as a MH admission within 365 days of a MHA detention spell that did not start as a new MHA detention.

The percentage of MHA detentions that had a subsequent MH admission within 12 months has increased in 2022/23 compared to earlier years (@tbl-readmissions-number). This also varies by ICB (@fig-readmissions-mh).

```{r}
#| label: tbl-readmissions-number
#| tbl-cap: Number of admissions within 12 months of a MHA detention.
tar_read(cyp_readmissions_perc) |>
  create_dt()
```

::: panel-tabset
##### Overview

```{r}
#| label: fig-readmissions-mh
#| fig-cap: Percentage of readmissions within 12 months.
tar_read(cyp_readmissions_perc_boxplot) +
  ggplot2::labs(title = "Percentage of readmissions within 12 months",
                subtitle = "England 2019/20 to 2022/23 by ICB.")
```

##### Summary

```{r}
#| label: tbl-readmissions-summary
#| tbl-cap: Summary of percentage of readmissions within 12 months.
tar_read(cyp_readmissions_perc_icb_code) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-readmissions
#| tbl-cap: Percentage of readmissions within 12 months by ICB.
tar_read(cyp_readmissions_perc_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 70, 9))
```
:::

### Travel distance

#### Overview of metric

The Ward Location Distance from Home variable represents the geographical distance between a patient's residential address and the ward location where they received treatment. This metric is commonly used to assess the accessibility of healthcare services and inform population health planning. By understanding the distance between a patient's home and the healthcare facility, we can gain insights into the convenience and proximity of services, identify potential barriers to healthcare access, and inform resource allocation decisions. Additionally, this data can be used for research studies to examine the relationship between travel distance and patient outcomes.

**Definition:** This variable represents the geographical distance between a patient's residential address and the ward location where they received treatment. It is often used to assess the accessibility of healthcare services and inform population health planning.

**Purpose:**

-   Accessibility: Measures the convenience and proximity of healthcare facilities for patients.

-   Population Health: Informs resource allocation and planning by understanding the distribution of healthcare services in relation to population density.

-   Research: Can be used to study the impact of travel distance on patient outcomes and adherence to treatment.

**Interpretation:**

-   Lower values: Indicate greater accessibility and convenience for patients.

-   Higher values: Suggest potential barriers to healthcare access, such as long travel times or transportation difficulties.

**Potential Use Cases:**

-   Patient Experience: Analyzing the relationship between ward location distance and patient satisfaction or outcomes.

-   Health Disparities: Identifying disparities in healthcare access based on geographic location.

-   Service Planning: Informing the location of new healthcare facilities or the expansion of existing services.

#### **Calculation of 'Ward location distance from home'**

Derived from: SITE CODE (OF TREATMENT)\
\
NHS Digital guidance: Look up the postcode for this SiteCode in the ODS reference data file that holds the address for organisation sites. Then identify the Northings\\Eastings for this postcode by using the ONS NHS Postcode Directory file.

**Reference:** <https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/data-set-technical-guidance/technical-glossary>

#### Travel distance

::: panel-tabset
#### Overview

::: panel-tabset
##### Chart

```{r}
#| label: fig-average-distance-chart
#| fig-cap: "Chart summarizing average distance per financial year."
tar_read(chart_DTT_FY)
```

##### Table

```{r}
#| label: fig-average-distance-fy
#| fig-cap: "Table summarizing average distance per financial year."
tar_read(table_DTT_FY) %>% create_dt()
```
:::

#### Gender Variation

::: panel-tabset
##### Chart

```{r}
#| label: fig-gender-average-distance-chart
#| fig-cap: "Chart summarizing average distance per gender and financial year."
tar_read(chart_DTT_gender_FY)
```

##### Table

```{r}
#| label: fig-gender-average-distance 
#| fig-cap: "Table summarizing average distance per gender and financial year."
tar_read(DTT_gender_FY) %>% create_dt()
```
:::

#### Age Group Variation

::: panel-tabset
##### Chart

```{r}
#| label: fig-age-group-average-distance-chart
#| fig-cap: "Chart summarizing average distance per age group and financial year."
tar_read(chart_DTT_age_group_FY)
```

##### Table

```{r}
#| label: fig-age-group-average-distance 
#| fig-cap: "Table summarizing average distance per age group and financial year."
tar_read(DTT_age_group_FY) %>% create_dt()
```
:::

#### Ethnicity Variation

::: panel-tabset
##### Chart

```{r}
#| label: fig-ethnic-category-average-distance-chart
#| fig-cap: "Chart summarizing average distance per ethnic categroy and financial year."
tar_read(chart_DTT_ethnic_FY)
```

##### Table

```{r}
#| label: fig-ethnic-average-distance 
#| fig-cap: "Table summarizing average distance per ethnic category and financial year."
tar_read(DTT_ethnic_FY) %>% create_dt()
```
:::

#### Socioeconomic variation

::: panel-tabset
##### Chart

```{r}
#| label: fig-IMD-average-distance-chart
#| fig-cap: "Chart summarizing average distance per IMD and financial year."
tar_read(chart_DTT_IMD_FY)
```

##### Table

```{r}
#| label: fig-imd-average-distance 
#| fig-cap: "Table summarizing average distance per IMD decile and financial year."
tar_read(DTT_IMD_FY) %>% create_dt()
```
:::
:::

#### Travel distance by number of admissions

::: panel-tabset
#### Overview

::: panel-tabset
##### Chart

```{r}
#| label: fig-comparison-admissions-distance
#| fig-cap: "Comparison of Total Admissions and Travel Distance Over Time."
tar_read(chart_admissions_vs_distance)
```

##### Table

```{r}
#| label: fig-DTT-FY-with-admissions-table
#| fig-cap: "Table showing average distance per admission and total admissions by financial year."

tar_read(table_DTT_FY_with_admissions)

```
:::

#### Gender

::: panel-tabset
##### Chart

```{r}
#| label: fig-gender-admissions-distance-chart
#| fig-cap: "Comparison of Total Admissions and Travel Distance by Gender Over Time."
tar_read(chart_DTT_gender_FY_with_admissions)
```

##### Table

```{r}
#| label: fig-gender-average-distance-admissions
#| fig-cap: "Table summarizing average distance and total admissions by gender and financial year."
tar_read(table_DTT_gender_FY_with_admissions)
```
:::

#### Age Group

::: panel-tabset
##### Chart

```{r}
#| label: fig-age-admissions-distance-chart
#| fig-cap: "Comparison of Total Admissions and Travel Distance by Age Group Over Time."
tar_read(chart_DTT_age_group_FY_with_admissions)
```

##### Table

```{r}
#| label: fig-age-group-average-distance-admissions
#| fig-cap: "Table summarizing average distance and total admissions by age group and financial year."
tar_read(table_DTT_age_group_FY_with_admissions)
```
:::

#### Ethnicity

::: panel-tabset
##### Chart

##### Table

```{r}
#| label: fig-ethnic-category-average-distance-admissions
#| fig-cap: "Table summarizing average distance and total admissions by ethnic category and financial year."
tar_read(table_DTT_ethnic_FY_with_admissions)
```
:::

#### IMD

::: panel-tabset
##### Chart

##### Table

```{r}
#| label: fig-IMD-average-distance-admissions
#| fig-cap: "Table summarizing average distance and total admissions by IMD and financial year."
tar_read(table_DTT_IMD_FY_with_admissions)
```
:::
:::

#### Questions for Andi

1.  FY - range?
