---
title: "Mental Health Act"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)

source("R/quarto_formatting.R")
```

## Introduction

## Data and methods

## Summary

## Detailed analysis

### Mental Health Act Detention Spells Length of Stay

This is the length of the MHA detention spell, where consecutive MHA detentions would be considered the same spell. 

In 2022/23, the distribution of MHA detention spells length of stay is right-skewed with the majority being 1 day long (@fig-los-histo). There is another peak in the distribution at about day 27, probably a result of detentions under section 2 of the MHA being limited to up to 28 days (@fig-los-histo-zoomed). 

There are some very long lengths of stay because ...

::: panel-tabset

#### All LoS

```{r}
#| label: fig-los-histo
#| fig-cap: Histogram of lengths of MHA detention spells for 2023-24.
tar_read(cyp_los_histo)
```

#### LoS < 50 days

```{r}
#| label: fig-los-histo-zoomed
#| fig-cap: Histogram of lengths of MHA detention spells less than 50 days for 2023-24.
tar_read(cyp_los_histo_zoomed)
```

:::

```{r}
los_median_23_24 <- tar_read(cyp_los_median) |>
  dplyr::filter(der_financial_year == "2023/24") |>
  dplyr::summarise(median = median(value)) |>
  dplyr::pull()
```


The median length of MHA detention spell has increased (@fig-los-mh) over time to `r los_median_23_24` days in 2023/24. This varies by ICB (@tbl-los-icb).

::: panel-tabset

##### Overview

```{r}
#| label: fig-los-mh
#| fig-cap: Median length of Mental Health Act detention spell.
tar_read(cyp_los_boxplot)  +
  ggplot2::labs(y = "Median length of MHA detention spell",
                title = "Median length of MHA detention spell",
                subtitle = "England 2019/20 to 2023/24 by ICB")
```

##### Summary

```{r}
#| label: tbl-los-summary
#| tbl-cap: Summary of median length of Mental Health Act detention spell.
tar_read(cyp_los_median) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-los-icb
#| tbl-cap: Median length of Mental Health Act detention spell by ICB.
tar_read(cyp_los_median_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 39, 5))
```

:::

Median length of stay for MHA detention spells are higher for females compared to males, higher to under 18s compared to 18-24 year olds and also show ethnic group and socioeconomic variation.

::: panel-tabset

##### Overview

::: panel-tabset

###### Chart

```{r}
#| label: fig-los-line
#| fig-cap: Median length of MHA detention over time.
tar_read(cyp_los_plot_overview)$plot
```

###### Table

```{r }
tar_read(cyp_los_plot_overview)$table |>
  create_dt()
```

:::

##### Gender variation

::: panel-tabset

###### Chart

```{r los_gender_plot}
#| label: fig-los-gender
#| fig-cap: Median length of MHA detention over time by gender.
tar_read(cyp_los_plot_gender)$plot
```

###### Table

```{r los_gender_table}
tar_read(cyp_los_plot_gender)$table |>
  create_dt()
```

:::

##### Age-group variation

::: panel-tabset

###### Chart

```{r los_age_plot}
#| label: fig-los-age
#| fig-cap: Median length of MHA detention over time by age group.
tar_read(cyp_los_plot_age_group)$plot
```

###### Table

```{r los_age_table}
tar_read(cyp_los_plot_age_group)$table |>
  create_dt()
```

:::

##### Ethnic group variation

::: panel-tabset

###### Chart

```{r los_ethnic_plot}
#| label: fig-los-ethnic
#| fig-cap: Median length of MHA detention over time by ethnic category.
tar_read(cyp_los_plot_ethnic_category)$plot
```

###### Table

```{r los_ethnic_table}
tar_read(cyp_los_plot_ethnic_category)$table |>
  create_dt()
```

:::

##### Socioeconomic variation

::: panel-tabset

###### Chart1

```{r los_imd_plot}
#| label: fig-los-imd
#| fig-cap: Median length of MHA detention over time by Index of Multiple Deprivation decile.
tar_read(cyp_los_plot_imd_decile)$plot
```

###### Table

```{r los_imd_table}
tar_read(cyp_los_plot_imd_decile)$table |>
  create_dt()
```

:::

:::


### Mental Health Act Re-detentions

The number of Mental Health Act (MHA) re-detentions within 12 months is defined as the number of subsequent MHA detention end-dates within 365 days of the first MHA detention end-date in a financial year.

For example, if a patient had the following MHA detention end dates:

2024-04-05\
2024-08-07\
2025-04-02\
2025-04-16\

Then the number of MHA re-detentions within 12 months for the financial years 2024/25 and 2025/26 would be 2 and 1, respectively. The percentage of MHA detentions that had a re-detention within 12 months for financial years 2024/25 and 2025/26 would be 100% and 50%, respectively.

```{r}
redet <- tar_read(cyp_redetentions) |>
  dplyr::rename(redetentions = attends)
```

About half of MHA detentions had a re-detention within 12 months (@tbl-redetentions-number) but this varies by ICB (@fig-redetentions-mh).

```{r}
#| label: tbl-redetentions-number
#| tbl-cap: Number of MHA detentions and redetentions within 12 months.
tar_read(cyp_redetentions) |>
  dplyr::summarise(detentions = sum(detentions),
                   redetentions = sum(attends),
                   .by = der_financial_year) |>
  mutate(perc = janitor::round_half_up(redetentions / detentions, 4)) |>
  create_dt()
```

::: panel-tabset

###### Overview

```{r}
#| label: fig-redetentions-mh
#| fig-cap: Percentage of MHA redetentions within 12 months.
tar_read(cyp_redetentions_perc_boxplot) +
  ggplot2::labs(title = "Percentage of redetentions within 12 months",
                subtitle = "England 2019/20 to 2022/23 by ICB.")
```

###### Summary

```{r}
#| label: tbl-redetentions-summary
#| tbl-cap: Summary of percentage of MHA redetentions within 12 months.
tar_read(cyp_redetentions_perc) |>
  get_quantiles_and_mean()
```

###### ICB Variation

```{r}
#| label: tbl-redetentions
#| tbl-cap: Percentage of MHA redetentions within 12 months by ICB.
tar_read(cyp_redetentions_perc_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 0.9, 0.125))
```
:::

#### Formal and Informal Mental Health Act Re-detentions

The majority or MHA re-detentions are formal with less than 1% being informal or unknown.

```{r}
cyp_redetentions_legal_status_perc <- tar_read(cyp_redetentions_legal_status) |>
  dplyr::summarise(redetentions = sum(attends), .by = c(legal_status, 
                                                         der_financial_year)) |>
  dplyr::mutate(perc = janitor::round_half_up(redetentions / sum(redetentions), 4), 
                .by = der_financial_year) 

cyp_redetentions_legal_status_perc |>
  dplyr::select(-perc) |>
  tidyr::pivot_wider(names_from = legal_status, values_from = redetentions) |>
  create_dt()
```

#### Re-admissions after a Mental Health Act detention

A re-admission is defined as a MH admission within 365 days of a MHA detention that did not start as a new MHA detention.

The percentage of MHA detentions that had a subsequent MH admission within 12 months has increased in 2022/23 compared to earlier years (@tbl-readmissions-number). This also varies by ICB (@fig-readmissions-mh).

```{r}
#| label: tbl-readmissions-number
#| tbl-cap: Number of admissions within 12 months of a MHA detention.
tar_read(cyp_readmissions) |>
  dplyr::summarise(detentions = sum(detentions),
                   readmissions = sum(attends),
                   .by = der_financial_year) |>
  mutate(perc = janitor::round_half_up(readmissions / detentions, 4)) |>
  create_dt()
```

::: panel-tabset

##### Overview

```{r}
#| label: fig-readmissions-mh
#| fig-cap: Percentage of readmissions within 12 months.
tar_read(cyp_readmissions_perc_boxplot) +
  ggplot2::labs(title = "Percentage of readmissions within 12 months",
                subtitle = "England 2019/20 to 2022/23 by ICB.")
```

##### Summary

```{r}
#| label: tbl-readmissions-summary
#| tbl-cap: Summary of percentage of readmissions within 12 months.
tar_read(cyp_readmissions_perc) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-readmissions
#| tbl-cap: Percentage of readmissions within 12 months by ICB.
tar_read(cyp_readmissions_perc_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23"), 
                 seq(0, 0.5, 0.07))
```

:::





### Travel distance

#### Overview of metric

The Ward Location Distance from Home variable represents the geographical distance between a patient's residential address and the ward location where they received treatment. This metric is commonly used to assess the accessibility of healthcare services and inform population health planning. By understanding the distance between a patient's home and the healthcare facility, we can gain insights into the convenience and proximity of services, identify potential barriers to healthcare access, and inform resource allocation decisions. Additionally, this data can be used for research studies to examine the relationship between travel distance and patient outcomes.

**Definition:** This variable represents the geographical distance between a patient's residential address and the ward location where they received treatment. It is often used to assess the accessibility of healthcare services and inform population health planning.

**Purpose:**

-   Accessibility: Measures the convenience and proximity of healthcare facilities for patients.

-   Population Health: Informs resource allocation and planning by understanding the distribution of healthcare services in relation to population density.

-   Research: Can be used to study the impact of travel distance on patient outcomes and adherence to treatment.

**Interpretation:**

-   Lower values: Indicate greater accessibility and convenience for patients.

-   Higher values: Suggest potential barriers to healthcare access, such as long travel times or transportation difficulties.

**Potential Use Cases:**

-   Patient Experience: Analyzing the relationship between ward location distance and patient satisfaction or outcomes.

-   Health Disparities: Identifying disparities in healthcare access based on geographic location.

-   Service Planning: Informing the location of new healthcare facilities or the expansion of existing services.

#### **Calculation of 'Ward location distance from home'**

Derived from: SITE CODE (OF TREATMENT)\
\
NHS Digital guidance: Look up the postcode for this SiteCode in the ODS reference data file that holds the address for organisation sites. Then identify the Northings\\Eastings for this postcode by using the ONS NHS Postcode Directory file.

**Reference:** <https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/data-set-technical-guidance/technical-glossary>

#### Summary

#### Variation

##### ICB

##### Gender

##### Age-group

##### Ethnic group

##### Socioeconomic variation
