---
title: "Mental Health attendances in urgent and emergency care settings"
author: "Andy Hood et al"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 3
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)
library(gt)

```

## Introduction

The care quality commission (CQC) are looking to supplement their State of Care (SOC) report with relevant additional data and analysis. The latest state of care report can be found here <https://www.cqc.org.uk/publications/major-report/state-care/2022-2023>.

One of the key areas of interest is around perceived 'system failure' as measured by (inappropriate/avoidable?) attendance to urgent and emergency care settings for people with mental health problems.

## Data and methods

We summarised routine data from the Emergency Care Dataset (ECDS), NHS 111 dataset and the Mental Health Services Dataset (MHSDS) via NCDR <https://www.ardengemcsu.nhs.uk/services/business-intelligence/ncdr/> according to specific population sub-groups and activity break-downs at Integrated Care Board (ICB) level. The period covers the 5 financial years from 1st April 2019 to 31st March 2024.

Patients were linked using common pseudonymised ID from ECDS and NHS111 to the MHSDS to establish whether they were known to specialist mental health services (referred and/or treated) or not.

The presence of a presenting mental health problem in **A&E data** was determined by all descendant SNOMED codes (n=1580) for 'Mental Health Disorders' in either the Chief Complaint field *OR* the 1st Diagnosis field. Codes relating to dementia, memory & cognition OR substance misuse were excluded.

The presence of mental health in **NHS 111 data** were established using a bespoke reference set for the symptom group comprising the specific mental health groups *AND* additional mentions of anxiety, depression, stress, suicidal thoughts or mental health in other symptom groups. Codes relating to dementia, memory & cognition AND substance misuse were excluded.

## Summary

-   Point1
-   Point2
-   Point3
-   Point4
-   Point5
-   Point6

## Accident and Emergency attendance

```{r ae_summary_bits, include = FALSE}

ae_all <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all)/1000000

ae_all_cost <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all_cost)/1000000000

ae_mh <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh)/1000

ae_mh_cost <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh_cost)/1000000

```

These attendances are those specified within the ECDS as 'Type 1' emergency department attendances and typically should be for those patients with life-threatening symptoms or injuries requiring immediate triage and treatment.

The average A&E attendance in England costs £280 with those conveyed by ambulance costing an additional £390 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

During 2023/24 there were `r round(ae_all,2)` million Type 1 attendances costing £ `r round(ae_all_cost,2)` billion. Of those attendances c`r round(ae_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(ae_mh_cost,2)` million.

```{r top_ten_causes}

gt_causes <- gt(tar_read(ae_diag) |> 
  filter(der_financial_year == '2023/24') |> 
  group_by(term) |> 
  summarise(attends = sum(attends)) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n=10)) |> 
  
  tab_header(
    title = "Top 10 SNOMED codes in A&E presentation",
    subtitle = "All of England, 2023/24"
  )

gt_causes

```

### Key metrics

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport.

```{r ae_arrive_mode}

tar_read(ae_trans_barplot)

```

#### Waiting times in ED

The ECDS allows calculation of times from arrival to treatment and from treatment to departure.

```{r ae_wait_plot}

tar_read(ae_times_plot)

```

#### Mental health 'frequent fliers'

```{r freq_fly_figs, include = FALSE}
avg_all_attends <- tar_read(ae_freq) |> summarise(total = sum(all_pats)) |> pull(total)/5

avg_mh_attend <- tar_read(ae_freq) |> summarise(total = sum(mh_pats)) |> pull(total)/5

perc_mh <- round(avg_mh_attend/avg_all_attends*100,1)

avg_ff_attend <- tar_read(ae_freq) |> summarise(total = sum(mh_freqfly)) |> pull(total)/5

perc_freq <- round(avg_ff_attend/avg_mh_attend*100,1)

```

Frequent fliers can be defined as those attending 5 or more times in a 12 month period with a primary complaint/diagnosis of mental health disorders.

Over the past 5 years in England, on average, there are `r comma(round(avg_mh_attend))` patients presenting with a mental health problem. Of these around `r comma(round(avg_ff_attend))` ( `r perc_freq`% ) are those who attend frequently. This varies by ICB area.

```{r ae_freq_icb_plot}

tar_read(ae_freq_boxplot)

```

### Detailed breakdowns

#### Geographical variation

```{r icb_map_allmh}
tar_read(icb_ed_map_2324)
```

#### Gender variation

#### Age-group variation

#### Ethnic group variation

#### Rural-urban variation

## Walk-in and Urgent Care attendance

These attendances are those specified within the ECDS as 'Type 3' urgent treatment centre and 'Type 4' walk-in centre attendances and should be typically self-conveyed patients requiring urgent assessment and treatment to prevent deterioration.

The average urgent care attendance in England costs £107 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

### Key metrics

#### Percentage of all attendances with MH as primary reason/diagnosis

Boxplot (@fig-perc-mh) suggests greater % of MH attends in 2023/24 over ICBs:

```{r}
#| label: "fig-perc-mh"
tar_read(uec_mh_attends_boxplot)
```

Looking at table (@tbl-perc) of ICBs over time:

-   there are some ICBs that have consistent % of MH attends over the financial years.
    -   For example Black Country and Herefordshire & Worcestershire have low %.
-   Sussex have high % compared to other ICBs.
-   Birmingham and Solihull is interesting, since it had a higher % in 2019/20, but low % for later fin years.

```{r}
#| label: "tbl-perc"
tar_read(uec_mh_attends_table)
```

#### Percentage of MH attendances known to specialist services

Boxplot (@fig-perc-mh-known) suggests lower % of MH attends were known in 2023/24 over ICBs:

```{r}
#| label: "fig-perc-mh-known"
tar_read(uec_mh_known_boxplot)
```

Looking at table (@tbl-perc-known) of ICBs over time:

-   Again Sussex have high % compared to other ICBs.

```{r}
#| label: "tbl-perc-known"
tar_read(uec_mh_known_table)
```

### Detailed breakdowns

#### Geographical variation

```{r icb_map_allmh_uec}
tar_read(icb_uec_map_2324)
```

#### Gender variation

#### Age-group variation

#### Ethnic group variation

#### Rural-urban variation

## NHS 111 calls

Callers to the NHS 111 service would be typically those unable to reach primary care services or to find out information and the appropriate treatment setting for a non-urgent health complaint.

The average cost of a call to NHS 111 is around £12 with those requiring clinical call-back an extra £15 and hear-treat or see-treat ambulance (999 referrals) costing on average another £203 <https://www.ncbi.nlm.nih.gov/books/NBK575169/>.

### Key metrics

### Detailed breakdowns

#### Geographical variation

#### Gender variation

#### Age-group variation

#### Ethnic group variation

#### Rural-urban variation
