---
title: "Mental Health attendances in urgent and emergency care settings"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)
library(gt)

source("R/quarto_formatting.R")
```

## Introduction

The care quality commission (CQC) are looking to supplement their State of Care (SOC) report with relevant additional data and analysis. The latest state of care report can be found here <https://www.cqc.org.uk/publications/major-report/state-care/2022-2023>.

One of the key areas of interest is around perceived 'system failure' as measured by (inappropriate/avoidable?) attendance to urgent and emergency care settings for people with mental health problems.

## Data and methods

We summarised routine data from the Emergency Care Dataset (ECDS), NHS 111 dataset and the Mental Health Services Dataset (MHSDS) via NCDR <https://www.ardengemcsu.nhs.uk/services/business-intelligence/ncdr/> according to specific population sub-groups and activity break-downs at Integrated Care Board (ICB) level. The period covers the 5 financial years from 1st April 2019 to 31st March 2024.

Patients were linked using common pseudonymised ID from ECDS and NHS111 to the MHSDS to establish whether they were known to specialist mental health services (referred and/or treated) or not.

The presence of a presenting mental health problem in **A&E data** was determined by all descendant SNOMED codes (n=1580) for 'Mental Health Disorders' in either the Chief Complaint field *OR* the 1st Diagnosis field. Codes relating to dementia, memory & cognition OR substance misuse were excluded.

The presence of mental health in **NHS 111 data** were established using a bespoke reference set for the symptom group comprising the specific mental health groups *AND* additional mentions of anxiety, depression, stress, suicidal thoughts or mental health in other symptom groups. Codes relating to dementia, memory & cognition AND substance misuse were excluded.

Denominator data on population numbers by sub-group (age, gender, imd decile, urban-rural) were sourced from ONS mid-year estimates at LSOA level grouped to 2023 ICB boundaries. As 2023 MYE has not yet been published for LSOA, the 2022 data has been rolled forward as a proxy. For ethnic category populations, the 2011 and 2021 census data has been used to impute linear estimates of the intervening and following years. Calendar year data has been fully attributed to the financial year in which it falls e.g. 2021 = 2021/22.

## Summary

-   Point1
-   Point2
-   Point3
-   Point4
-   Point5
-   Point6

## Accident and Emergency attendance

```{r ae_summary_bits, include = FALSE}

ae_all <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all)/1000000

ae_all_cost <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all_cost)/1000000000

ae_mh <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh)/1000

ae_mh_cost <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh_cost)/1000000

```

These attendances are those specified within the ECDS as 'Type 1' emergency department attendances and typically should be for those patients with life-threatening symptoms or injuries requiring immediate triage and treatment.

The average A&E attendance in England costs £280 with those conveyed by ambulance costing an additional £390 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

During 2023/24 there were `r round(ae_all,2)` million Type 1 attendances costing £ `r round(ae_all_cost,2)` billion. Of those attendances c`r round(ae_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(ae_mh_cost,2)` million.

```{r ae_top_ten_causes}

gt_causes <- gt(tar_read(ae_diag) |> 
  filter(der_financial_year == '2023/24', ec_department_type == 1) |> 
  group_by(term) |> 
  summarise(attends = sum(attends)) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n=10)) |> 
  
  tab_header(
    title = "Top 10 SNOMED codes in A&E presentation",
    subtitle = "All of England, 2023/24"
  )

gt_causes

```

### Key metrics

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport. Breakdowns by population demographics can be seen in @sec-arrival-mode-breakdowns.

::: panel-tabset
##### All modes
```{r ae_arrive_mode}
tar_read(ae_trans_barplot)
```

##### Trend
```{r ae_arrive_mode_trend}
tar_read(ae_trans_trends)
```

##### Table
```{r ae_arrive_mode_table}
tar_read(ae_transp_trends)
```
:::

#### Time of arrival

Patients presenting with mental health complaints are more likely to attend after 8pm and before 6am than their contemporaries with physical health complaints - this could be suggestive of lesser access to out-of-hours psychiatric and crisis support or the increased likelihood of crisis occurring when people are alone.

::: panel-tabset
##### Chart
```{r ae_arrive_time_plot}
tar_read(ae_toa_plot)
```

##### Table
```{r ae_arrive_time_table}
tar_read(ae_toa_summary)
```
:::

#### Leaving the department

SNOMED departure status codes allow us to identify patients that left the department before either assessment or treatment was completed.

::: panel-tabset
##### Chart
```{r ae_left_plot}
tar_read(ed_left_chart)
```

##### Table
```{r ae_left_summary}
tar_read(ae_left_summary)
```
:::


#### Waiting times in ED

The ECDS allows calculation of times from arrival to assessment, treatment, conclusion and to departure.

::: panel-tabset
##### Assessment
```{r ae_wait_assess_plot}
tar_read(ae_times_assess_plot)
```

##### Treatment
```{r ae_wait_treat_plot}
tar_read(ae_times_treat_plot)
```

##### Conclusion
```{r ae_wait_conclude_plot}
tar_read(ae_times_conclude_plot)
```

##### Departure
```{r ae_wait_depart_plot}
tar_read(ae_times_depart_plot)
```
:::

#### Mental health 'frequent fliers'

```{r freq_fly_figs, include = FALSE}
avg_all_attends <- tar_read(ae_freq) |> summarise(total = sum(all_pats)) |> pull(total)/5

avg_mh_attend <- tar_read(ae_freq) |> summarise(total = sum(mh_pats)) |> pull(total)/5

perc_mh <- round(avg_mh_attend/avg_all_attends*100,1)

avg_ff_attend <- tar_read(ae_freq) |> summarise(total = sum(mh_freqfly)) |> pull(total)/5

perc_freq <- round(avg_ff_attend/avg_mh_attend*100,1)

```

Frequent fliers can be defined as those attending 5 or more times in a 12 month period with a primary complaint/diagnosis of mental health disorders.

Over the past 5 years in England, on average, there are `r comma(round(avg_mh_attend))` patients presenting with a mental health problem. Of these around `r comma(round(avg_ff_attend))` ( `r perc_freq`% ) are those who attend frequently. This varies by ICB area.

```{r ae_freq_icb_plot}

tar_read(ae_freq_boxplot)

```

### Detailed breakdowns

#### Percentage of all Type 1 attendances with Mental Health as primary reason/diagnosis {#sec-type1-mh-attends}

@fig-type1-perc-mh suggests the median percentage of Type 1 attendances with Mental Health as the primary reason/diagnosis is comparable with pre-pandemic levels. Breakdowns by ICB are in @tbl-type1-perc and breakdowns by population demographics below. 

::: panel-tabset

###### Overview

```{r}
#| label: "fig-type1-perc-mh"
#| fig-cap: "Percentage of Mental Health Type 1 attendances over time."
tar_read(type1_mh_attends_boxplot)
```

###### ICB Variation

```{r}
#| label: "tbl-type1-perc"
#| tbl-cap: "Percentage of Mental Health Type 1 attendances by ICB."
tar_read(type1_mh_attends_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 2.5, 0.35))
```

:::

Breakdowns for the number of Type 1 attendances with MH as primary reason/diagnosis are below.

##### Geographical variation

```{r icb_map_allmh}
tar_read(icb_ed_map_2324)
```

::: panel-tabset

##### Gender variation

:::: panel-tabset

###### Chart

```{r type1_attends_gender_plot}
#| label: "fig-type1_attends-gender"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by gender."
tar_read(gender_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_gender_table}
tar_read(gender_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```

::::

##### Age-group variation

:::: panel-tabset

###### Chart

```{r type1_attends_age_plot}
#| label: "fig-type1_attends-age"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by age group."
tar_read(age_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_age_table}
tar_read(age_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```

::::

##### Ethnic group variation

:::: panel-tabset

###### Chart

```{r type1_attends_ethnic_plot}
#| label: "fig-type1_attends-ethnic"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by ethnic category."
tar_read(ethnic_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_ethnic_table}
tar_read(ethnic_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```

::::

##### Socioeconomic variation

:::: panel-tabset

###### Chart1

```{r type1_attends_imd_plot}
#| label: "fig-type1_attends-imd"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by Index of Multiple Deprivation decile."
tar_read(imd_plot)$type1_mh_attends
```

###### Chart2

```{r type1_attends_imd_plot2}
tar_read(ae_attends_imd)
```

###### Table

```{r type1_attends_imd_table}
tar_read(imd_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```

::::

##### Rural-urban variation

:::: panel-tabset

###### Chart

```{r type1_attends_rural_plot}
#| label: "fig-type1_attends-rural"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by rural and urban populations."
tar_read(rural_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_rural_table}
tar_read(rural_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```

::::

:::

#### Percentage of MH Type 1 attendances known to specialist services {#sec-type1-mh-known}

The percentages of Mental Health Type 1 attendances where the patient is already known to specialist services varies over time (@fig-type1-mh-known-boxplot) and by ICB (@tbl-type1-mh-known). 

::: panel-tabset

###### Overview

```{r}
#| label: "fig-type1-mh-known-boxplot"
#| fig-cap: "Percentage of Mental Health Type 1 attendances known to specialist services over time."
tar_read(type1_mh_known_boxplot)
```

###### ICB Variation

```{r}
#| label: "tbl-type1-mh-known"
#| tbl-cap: "Percentage of Mental Health Type 1 attendances known to specialist services by ICB."
tar_read(type1_mh_known_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(60, 90, 4)) 
```

:::

Breakdowns for the number of Type 1 attendances with MH as primary reason/diagnosis where the patient was known to specialist services are below.

::: panel-tabset

##### Gender variation

:::: panel-tabset

###### Chart

```{r type1_known_gender_plot}
#| label: "fig-type1_known-gender"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances where the patient was known to specialist services by gender."
tar_read(gender_plot)$type1_mh_known
```

###### Table

```{r type1_known_gender_table}
tar_read(gender_breakdowns)$type1_mh_known |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```

::::

##### Age-group variation

:::: panel-tabset

###### Chart

```{r type1_known_age_plot}
#| label: "fig-type1_known-age"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances where the patient was known to specialist services by age group."
tar_read(age_plot)$type1_mh_known
```

###### Table

```{r type1_known_age_table}
tar_read(age_breakdowns)$type1_mh_known |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```

::::

##### Ethnic group variation

:::: panel-tabset

###### Chart

```{r type1_known_ethnic_plot}
#| label: "fig-type1_known-ethnic"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances where the patient was known to specialist services by ethnic category."
tar_read(ethnic_plot)$type1_mh_known
```

###### Table

```{r type1_known_ethnic_table}
tar_read(ethnic_breakdowns)$type1_mh_known |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```

::::

##### Socioeconomic variation

:::: panel-tabset

###### Chart

```{r type1_known_imd_plot}
#| label: "fig-type1_known-imd"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances where the patient was known to specialist services by Index of Multiple Deprivation decile."
tar_read(imd_plot)$type1_mh_known
```

###### Table

```{r type1_known_imd_table}
tar_read(imd_breakdowns)$type1_mh_known |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```

::::

##### Rural-urban variation

:::: panel-tabset

###### Chart

```{r type1_known_rural_plot}
#| label: "fig-type1_known-rural"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances where the patient was known to specialist services by rural and urban populations."
tar_read(rural_plot)$type1_mh_known
```

###### Table

```{r type1_known_rural_table}
tar_read(rural_breakdowns)$type1_mh_known |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```

::::

:::

#### Arrival mode breakdowns {#sec-arrival-mode-breakdowns}

The breakdowns below are for the number of mental health attendances only.

::: panel-tabset

##### Gender variation

::: panel-tabset

###### Chart

```{r type1_arrival_mode_gender_plot}
#| label: "fig-arrival-mode-gender"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by arrival mode by gender."
tar_read(gender_plot)$type1_arrival_mode |> 
  modify_arrival_mode_plots()
```

###### Table

```{r type1_arrival_mode_gender_table}
tar_read(gender_breakdowns)$type1_arrival_mode |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```

:::

##### Age-group variation

:::: panel-tabset

###### Chart

```{r type1_arrival_mode_age_plot}
#| label: "fig-arrival-mode-age"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by arrival mode by age group."
tar_read(age_plot)$type1_arrival_mode |> 
  modify_arrival_mode_plots()
```

###### Table

```{r type1_arrival_mode_age_table}
tar_read(age_breakdowns)$type1_arrival_mode |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```

::::

##### Ethnic group variation

:::: panel-tabset

###### Chart

```{r type1_arrival_mode_ethnic_plot}
#| label: "fig-arrival-mode-ethnic"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by arrival mode by ethnic category."
tar_read(ethnic_plot)$type1_arrival_mode |> 
  modify_arrival_mode_plots()
```

###### Table

```{r type1_arrival_mode_ethnic_table}
tar_read(ethnic_breakdowns)$type1_arrival_mode |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```

::::

##### Socioeconomic variation

:::: panel-tabset

###### Chart

```{r type1_arrival_mode_imd_plot}
#| label: "fig-arrival-mode-imd"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by arrival mode by Index of Multiple Deprivation decile."
tar_read(imd_plot)$type1_arrival_mode |> 
  modify_arrival_mode_plots()
```

###### Table

```{r type1_arrival_mode_imd_table}
tar_read(imd_breakdowns)$type1_arrival_mode |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```

::::

##### Rural-urban variation

:::: panel-tabset

###### Chart

```{r type1_arrival_mode_rural_plot}
#| label: "fig-arrival-mode-rural"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health Type 1 attendances by arrival mode by rural or urban populations."
tar_read(rural_plot)$type1_arrival_mode |> 
  modify_arrival_mode_plots()
```

###### Table

```{r type1_arrival_mode_rural_table}
tar_read(rural_breakdowns)$type1_arrival_mode |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```

::::

:::

## Walk-in and Urgent Care attendance

These attendances are those specified within the ECDS as 'Type 3' urgent treatment centre and 'Type 4' walk-in centre attendances and should be typically self-conveyed patients requiring urgent assessment and treatment to prevent deterioration.

The average urgent care attendance in England costs £107 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

```{r uec_top_ten_causes}

gt_causes <- gt(tar_read(ae_diag) |> 
  filter(der_financial_year == '2023/24', ec_department_type %in% c(3,4)) |> 
  group_by(term) |> 
  summarise(attends = sum(attends)) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n=10)) |> 
  
  tab_header(
    title = "Top 10 SNOMED codes in UCC and WIC presentation",
    subtitle = "All of England, 2023/24"
  )

gt_causes

```

### Key metrics

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport. Breakdowns by population demographics can be seen in @sec-arrival-mode-breakdowns.

::: panel-tabset
##### All modes
```{r uec_arrive_mode}
tar_read()
```

##### Trend
```{r uec_arrive_mode_trend}
tar_read()
```

##### Table
```{r uec_arrive_mode_table}
tar_read()
```
:::

#### Time of arrival


::: panel-tabset
##### Chart
```{r uec_arrive_time_plot}
tar_read()
```

##### Table
```{r uec_arrive_time_table}
tar_read()
```
:::

#### Leaving the department

SNOMED departure status codes allow us to identify patients that left the department before either assessment or treatment was completed.

::: panel-tabset
##### Chart
```{r uec_left_plot}
tar_read()
```

##### Table
```{r uec_left_summary}
tar_read()
```
:::


### Detailed breakdowns

#### Percentage of all UEC attendances with MH as primary reason/diagnosis {#sec-uec-mh-attends}

@fig-uec-perc-mh suggests greater percentage of Mental Health UEC attendances in 2023/24. This varies by ICB (@tbl-uec-perc) where
some ICBs have consistently low or high percentages over time. 

::: panel-tabset

###### Overview

```{r}
#| label: "fig-uec-perc-mh"
#| fig-cap: "Percentage of Mental Health UEC attendances over time."
tar_read(uec_mh_attends_boxplot)
```

###### ICB Variation

```{r}
#| label: "tbl-uec-perc"
#| tbl-cap: "Percentage of Mental Health UEC attendances by ICB."
tar_read(uec_mh_attends_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 1.8, 0.25))
```

:::

Breakdowns for the number of attendances to UEC with MH as primary reason/diagnosis are below.

##### Geographical variation

```{r icb_map_allmh_uec}
tar_read(icb_uec_map_2324)
```

::: panel-tabset

##### Gender variation

:::: panel-tabset

###### Chart

```{r uec_attends_gender_plot}
#| label: "fig-uec-attends-gender"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances by gender."
tar_read(gender_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_gender_table}
tar_read(gender_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```

::::

##### Age-group variation

:::: panel-tabset

###### Chart

```{r uec_attends_age_plot}
#| label: "fig-uec-attends-age"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances by age group."
tar_read(age_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_age_table}
tar_read(age_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```

::::

##### Ethnic group variation

:::: panel-tabset

###### Chart

```{r uec_attends_ethnic_plot}
#| label: "fig-uec-attends-ethnic"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances by ethnic category."
tar_read(ethnic_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_ethnic_table}
tar_read(ethnic_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```

::::

##### Socioeconomic variation

:::: panel-tabset

###### Chart1

```{r uec_attends_imd_plot}
#| label: "fig-uec-attends-imd"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances by Index of Multiple Deprivation decile."
tar_read(imd_plot)$uec_mh_attends
```

###### Chart2

```{r uec_attends_imd_plot2}
tar_read(uec_attends_imd)
```


###### Table

```{r uec_attends_imd_table}
tar_read(imd_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```

::::

##### Rural-urban variation

:::: panel-tabset

###### Chart

```{r uec_attends_rural_plot}
tar_read(rural_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_rural_table}
#| label: "fig-uec-attends-rural"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances by rural and urban populations."
tar_read(rural_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```

::::

:::

#### Percentage of MH UEC attendances known to specialist services {#sec-uec-mh-known}

@fig-uec-mh-known-boxplot suggests lower percentage of Mental Health UEC attends were known in 2023/24. This again varies by ICB (@tbl-uec-mh-known).

::: panel-tabset

###### Overview

```{r}
#| label: "fig-uec-mh-known-boxplot"
#| fig-cap: "Percentage of Mental Health UEC attendances known to specialist services over time."
tar_read(uec_mh_known_boxplot)
```

###### ICB Variation

```{r}
#| label: "tbl-uec-mh-known"
#| tbl-cap: "Percentage of Mental Health UEC attendances known to specialist services by ICB."
tar_read(uec_mh_known_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(30, 90, 8)) 
```

:::

Breakdowns for the number of UEC attendances with MH as primary reason/diagnosis where the patient was known to specialist services are below.

::: panel-tabset

##### Gender variation

:::: panel-tabset

###### Chart

```{r uec_known_gender_plot}
#| label: "fig-uec_known-gender"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances where the patient was known to specialist services by gender."
tar_read(gender_plot)$uec_mh_known
```

###### Table

```{r uec_known_gender_table}
tar_read(gender_breakdowns)$uec_mh_known |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```

::::

##### Age-group variation

:::: panel-tabset

###### Chart

```{r uec_known_age_plot}
#| label: "fig-uec_known-age"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances where the patient was known to specialist services by age group."
tar_read(age_plot)$uec_mh_known
```

###### Table

```{r uec_known_age_table}
tar_read(age_breakdowns)$uec_mh_known |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```

::::

##### Ethnic group variation

:::: panel-tabset

###### Chart

```{r uec_known_ethnic_plot}
#| label: "fig-uec_known-ethnic"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances where the patient was known to specialist services by ethnic category."
tar_read(ethnic_plot)$uec_mh_known
```

###### Table

```{r uec_known_ethnic_table}
tar_read(ethnic_breakdowns)$uec_mh_known |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```

::::

##### Socioeconomic variation

:::: panel-tabset

###### Chart

```{r uec_known_imd_plot}
#| label: "fig-uec_known-imd"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances where the patient was known to specialist services by Index of Multiple Deprivation decile."
tar_read(imd_plot)$uec_mh_known
```

###### Table

```{r uec_known_imd_table}
tar_read(imd_breakdowns)$uec_mh_known |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```

::::

##### Rural-urban variation

:::: panel-tabset

###### Chart

```{r uec_known_rural_plot}
#| label: "fig-uec_known-rural"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health UEC attendances where the patient was known to specialist services by rural and urban populations."
tar_read(rural_plot)$uec_mh_known
```

###### Table

```{r uec_known_rural_table}
tar_read(rural_breakdowns)$uec_mh_known |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```

::::

:::

## NHS 111 calls

Callers to the NHS 111 service would be typically those unable to reach primary care services or to find out information and the appropriate treatment setting for a non-urgent health complaint.

The average cost of a call to NHS 111 is around £12 with those requiring clinical call-back an extra £15 and hear-treat or see-treat ambulance (999 referrals) costing on average another £203 <https://www.ncbi.nlm.nih.gov/books/NBK575169/>.

### Key metrics

### Detailed breakdowns

#### % of all calls being MH-related

@fig-nhs111-perc-mh suggests ... Breakdowns by ICB are in @tbl-nhs111-perc and breakdowns by population demographics can be seen below. 

::: panel-tabset

###### Overview

```{r}
#| label: "fig-nhs111-perc-mh"
#| fig-cap: "Percentage of Mental Health calls over time."
tar_read(nhs111_mh_calls_boxplot)
```

###### ICB Variation

```{r}
#| label: "tbl-nhs111-perc"
#| tbl-cap: "Percentage of Mental Health Type 1 calls by ICB."
tar_read(nhs111_mh_calls_table) |>
  color_gradient(c("2017/18", "2018/19", "2019/20", "2020/21", "2021/22"), 
                 seq(0, 1.8, 0.25))
```

:::

#### Geographical variation

::: panel-tabset

##### Gender variation

:::: panel-tabset

###### Chart

```{r nhs111_calls_gender_plot}
#| label: "fig-nhs111_calls-gender"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health related calls to NHS111 by gender."
tar_read(gender_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_gender_table}
tar_read(gender_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```

::::

##### Age-group variation

:::: panel-tabset

###### Chart

```{r nhs111_calls_age_plot}
#| label: "fig-nhs111_calls-age"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health related calls to NHS111 by age group."
tar_read(age_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_age_table}
tar_read(age_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```

::::

##### Socioeconomic variation

:::: panel-tabset

###### Chart1

```{r nhs111_calls_imd_plot}
#| label: "fig-nhs111_calls-imd"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health related calls to NHS111 by Index of Multiple Deprivation decile."
tar_read(imd_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_imd_table}
tar_read(imd_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```

::::

##### Rural-urban variation

:::: panel-tabset

###### Chart

```{r nhs111_calls_rural_plot}
#| label: "fig-nhs111_calls-rural"
#| fig-cap: "Rates per 100,000 population of the number of Mental Health related calls to NHS111 by rural and urban populations."
tar_read(rural_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_rural_table}
tar_read(rural_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```

::::

:::
