---
title: "Mental Health attendances in urgent and emergency care settings"
author: "Andy Hood, Anya Ferguson, Andre Bilbrough, Heather Humphreys"
format:
  html:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fig-format: svg
    mainfont: arial
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Summary
    toc-depth: 4
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 0px
      gutter-width: 1.5rem
editor: visual
execute:
  echo: false
  message: false
  warning: false
  cache: false
lang: en-GB
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
css: styles.css
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)

source("R/quarto_formatting.R")
```

## Introduction

The care quality commission (CQC) are looking to supplement their State of Care (SOC) report with relevant additional data and analysis. The latest state of care report can be found here <https://www.cqc.org.uk/publications/major-report/state-care/2022-2023>.

One of the key areas of interest is around perceived 'system failure' as measured by inappropriate or avoidable attendance to urgent and emergency care settings for people with mental health problems.

## Data and methods

We summarised routine data from the Emergency Care Dataset (ECDS), NHS 111 dataset and the Mental Health Services Dataset (MHSDS) via NCDR <https://www.ardengemcsu.nhs.uk/services/business-intelligence/ncdr/> according to specific population sub-groups and activity break-downs at Integrated Care Board (ICB) level. The period covers the 5 financial years from 1st April 2019 to 31st March 2024 unless otherwise stated.

Patients were linked using common pseudonymised ID from ECDS and NHS111 to the MHSDS to establish whether they were known to mental health services (referred and/or treated) or not.

The presence of a presenting mental health problem in **A&E data** was determined by all descendant SNOMED codes (n=1580) for 'Mental Health Disorders' in either the Chief Complaint field *OR* the 1st Diagnosis field. Codes relating to dementia, memory & cognition OR substance misuse were excluded.

The presence of mental health in **NHS 111 data** were established using a bespoke reference set for the symptom group comprising the specific mental health groups *AND* additional mentions of anxiety, depression, stress, suicidal thoughts or mental health in other symptom groups. Codes relating to dementia, memory & cognition AND substance misuse were excluded.

Denominator data on population numbers by sub-group (age, gender, imd decile, urban-rural) were sourced from ONS mid-year estimates at LSOA level grouped to 2023 ICB boundaries. As 2023 MYE has not yet been published for LSOA, the 2022 data has been rolled forward as a proxy. For ethnic category populations, the 2011 and 2021 census data has been used to impute linear estimates of the intervening and following years. In order to match 2001 categories used in SUS data <https://www.datadictionary.nhs.uk/data_elements/ethnic_category.html>, the Chinese category has been moved from 'Asian' to 'Other' in the 2011 and 2021 census population data <https://www.ons.gov.uk/census/census2021dictionary/variablesbytopic/ethnicgroupnationalidentitylanguageandreligionvariablescensus2021/ethnicgroup/classifications>.

In the detailed breakdowns sections, utilisation rates are presented by population sub-groups. These plots include a dashed black line for the average of the sub-groups (for example @fig-type1_attends-gender). A separate overview plot is provided to show the overall rates (for example @fig-type1-line).

Calendar year data has been fully attributed to the financial year in which it falls e.g. 2021 = 2021/22.

### Data quality notes for A&E attendances:

Chief complaint code (SNOMED): The current completion rate is 83% with 81% valid (correct/usable format) for 2024/25 which has improved from 78% complete and 75% valid in 2019/2020. The missing data in this field is more likely to represent more acute or complex conditions that are not self-reported or determinable prior to assessment.

Ethnic category codes: The current field completion is 98.2% overall with a validity rate (correct/usable format) of 91.5% for 2024/25; however there is significant variation in subgroups - for example range is 57% - 95% for some categories. Missing data is more likely to be from minority ethnic groups or where English is not first language therefore some of our rate calculations may understate the utilisation in these population groups.

## Summary

The likelihood of someone with a mental health problem contacting urgent and emergency care services varies greatly according to who they are and where they live. Females, those in adolescence, those in 'other' minority ethnic groups and those in deprived areas all access services at higher rates.

Attendance rates to Type 1 units (emergency departments) are falling gradually over time but those to urgent treatment centres are increasing. Patients with mental health difficulties also access services differently -- use ambulances more, arrive out of hours -- and are also more likely to wait longer in A&E departments for assessment and treatment than those with physical health complaints.

While the majority of these patients are known to mental health providers, a substantial minority are not. Typically, the most common presenting conditions are depression and anxiety though material numbers do have more severe disorders.

### Type 1 A&E attendances:

-   In 2023/24 there were around 16 million attendances to ED, of which 200,000 had a primary diagnosis of mental health condition. There is a 4-fold difference in % of attendances being MH specific across the 42 ICB areas.
-   These MH attendances cost \> £31m (not including ambulatory costs).
-   Depression and anxiety account for around 70% of MH attendances. Psychosis, Personality Disorder, Schizophrenia and Bipolar disorder attendances make up around 17% of MH presentations and eating disorders around 2%.
-   Typically, over 80% of these patients presenting in A&E are known to mental health services. This also means that nearly 20% are not.
-   Significantly more patients with MH disorders are conveyed to A&E by ambulance than those with physical ailments -- 40% vs 30%.
-   Patients presenting with MH issues are much more likely to attend between the hours of 8pm and 6am than those with physical issues.
-   Fewer patients with MH complaints (4.3%) leave the department before their assessment or treatment is concluded than those with physical complaints (5.7%).
-   In 2023/24, those presenting with MH issues waited longer in the department for assessment, treatment and to discharge from A&E. The average ED 'pathway' was almost an hour longer for MH patients.
-   There were almost 2,500 patients in England attending Emergency OR Urgent care departments 5 or more times during the latest 12 month period with a primary mental health complaint.
-   The rate of attendance in females (383/100,000 population) is significantly higher than males (309).
-   The adolescent/transitional age group, 18-21 years, consistently has the highest rates of presentation for MH. The 0-17 age group is the only one where the rates have increased since 2019/20.
-   Patients with MH issues from 'Other' ethnic categories (Chinese, Arab and other) present more frequently than other groups. Rates in the Mixed ethnic groups are increasing over time.
-   There is a very clear socio-economic gradient in MH presentation rates. IMD1 (most deprived) have rates over 3 times higher than IMD10 the least deprived. Similarly, patients from urban areas are around twice as likely to present in A&E with MH problems.

### Walk-in and UTC attendances:

-   In 2023/24 there were around 6.5 million attendances to Type 3 or 4 departments, of which 20,000 had a primary diagnosis of mental health condition.
-   These MH attendances cost £1.9m (not including ambulatory costs).
-   Depression and anxiety account for over 85% of MH attendances in this setting.
-   Typically, around 50% of these patients presenting in urgent care are known to mental health services, suggesting their mental health needs are less severe than those attending Type 1 Emergency Departments (80%+ known).
-   The overwhelming majority (90%+) of both MH and non-MH patients convey themselves to the urgent care departments.
-   Patients presenting with MH issues are slightly more likely to attend through the afternoon, evening and night than those with physical issues.
-   Fewer patients with MH complaints (2.3%) leave the department before their assessment or treatment is concluded than those with physical complaints (3.6%).
-   Generally speaking rates of MH presentation to Urgent Care or Walk-in Centres are increasing over time.
-   The rate of attendance in females (40/100,000 population) is significantly higher than males (32).
-   The adolescent/transitional age group, 18-21 years, consistently has the highest rates of presentation for MH.
-   Patients with MH issues from 'Other' ethnic categories (Chinese, Arab and other) present more frequently than other groups.
-   There is a very clear socio-economic gradient in MH presentation rates. IMD1 (most deprived) have rates over 3 times higher than IMD10 the least deprived. Similarly, patients from urban areas are nearly twice as likely to present in A&E with MH problems.

### Calls to NHS 111:

-   In 2023/24 there were around 33 million calls to NHS 111 services across England, of which 220,000 had a primary reason of mental health condition.
-   These MH specific calls cost £2.7m (not including ambulatory costs).
-   The 5 most common calls are related to worsening of known mental health issues. Suicide-related symptoms account for around 30% of all calls.
-   Around 85% of these patients calling NHS 111 with mental health issues are known to mental health services.
-   Patients presenting with MH issues are slightly more likely to call in the early morning and early evening.
-   More than 1 in 3 patients calling NHS 111 (34%) with primary mental health complaint are referred on to A&E department compared with only 11% with a physical complaint and this disparity is growing.
-   The rate of NHS 111 calls from females (430/100,000 population) is significantly higher than males (350).
-   The adolescent/transitional age group, 18-21 years, consistently has the highest rates of calls for MH.
-   There is a clear but reasonably shallow socio-economic gradient in MH related calls. IMD1 (most deprived) have rates just under 2 times higher than IMD10 the least deprived. Similarly, patients from urban areas are much more likely to call NHS 111 with MH problems.

## Detailed analysis - Accident and Emergency attendance {#sec-type1-overview}

```{r ae_summary_bits, include = FALSE}

ae_all <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all)/1000000

ae_all_cost <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(all_cost)/1000000000

ae_mh <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh)/1000

ae_mh_cost <- tar_read(ae_summary) |> filter(der_financial_year == '2023/24') |> pull(mh_cost)/1000000

```

These attendances are those specified within the ECDS as 'Type 1' emergency department attendances and typically should be for those patients with life-threatening symptoms or injuries requiring immediate triage and treatment.

The average A&E attendance in England costs £280 with those conveyed by ambulance costing an additional £390 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

During 2023/24 there were `r round(ae_all,2)` million Type 1 attendances costing £ `r round(ae_all_cost,2)` billion. Of those attendances c`r round(ae_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(ae_mh_cost,2)` million.

```{r}
#| label: tbl-type1-snomed
#| tbl-cap: Top 10 SNOMED codes in A&E presentation. All of England, 2023/24.

gt_causes <- tar_read(ae_diag) |> 
  filter(der_financial_year == '2023/24', ec_department_type == 1) |> 
  summarise(attends = sum(attends), .by = term) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n = 10) 

create_dt(gt_causes) 
```

### Access and patient pathway {#sec-type1-pathway}

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport.

::: panel-tabset
##### All modes

```{r}
#| label: fig-type1-arrival
#| fig-cap: Arrival mode to Type 1 A&E. All of England, 2023/24.
tar_read(ae_trans_barplot)
```

##### Trend

```{r}
#| label: fig-type1-arrival-trend
#| fig-cap: Arrival mode to Type 1 A&E. All of England, 2019/20 to 2023/24.
tar_read(ae_trans_trends)
```

##### Table

```{r}
#| label: tbl-type1-arrival-trend
#| tbl-cap: Arrival mode to Type 1 A&E. All of England, 2019/20 to 2023/24.
tar_read(ae_transp_trends) |>
  create_dt()
```
:::

#### Time of arrival

Patients presenting with mental health complaints are more likely to attend after 8pm and before 6am than their contemporaries with physical health complaints - this could be suggestive of lesser access to out-of-hours psychiatric and crisis support or the increased likelihood of crisis occurring when people are alone.

::: panel-tabset
##### Chart

```{r ae_arrive_time_plot}
#| label: fig-type1-toa
#| fig-cap: Hour of arrival to Type 1 A&E. All of England, 2023/24.
tar_read(ae_toa_plot)
```

##### Table

```{r ae_arrive_time_table}
#| label: tbl-type1-toa
#| tbl-cap: Number and percentage of attendances by time of arrival and financial year.
tar_read(ae_toa_summary) |>
  create_dt()
```
:::

#### Leaving the department

SNOMED departure status codes allow us to identify patients that left the department before either assessment or treatment was completed.

::: panel-tabset
##### Chart

```{r ae_left_plot}
#| label: fig-type1-left
#| fig-cap: Left department before assessment or treatment complete. All of England, 2023/24.
tar_read(ed_left_chart)
```

##### Table

```{r ae_left_summary}
#| label: tbl-type1-leave
#| tbl-cap: Number of attendances and patients leaving the departments before assessment or treatment by financial year.
tar_read(ae_left_summary) |>
  create_dt()
```
:::

#### Waiting times in ED

The ECDS allows calculation of times from arrival to assessment, treatment, conclusion and to departure.

::: panel-tabset
##### Assessment

```{r ae_wait_assess_plot}
#| label: fig-type1-waiting-ass
#| fig-cap: Waiting time for assessment, Type 1 A&E. All of England, 2023/24.
tar_read(ae_times_assess_plot)
```

##### Treatment

```{r ae_wait_treat_plot}
#| label: fig-type1-waiting-treat
#| fig-cap: Waiting time for treatment, Type 1 A&E. All of England, 2023/24.
tar_read(ae_times_treat_plot)
```

##### Conclusion

```{r ae_wait_conclude_plot}
#| label: fig-type1-waiting-concl
#| fig-cap: Waiting time for conclusion, Type 1 A&E. All of England, 2023/24.
tar_read(ae_times_conclude_plot)
```

##### Departure

```{r ae_wait_depart_plot}
#| label: fig-type1-waiting-depart
#| fig-cap: Waiting time to leave department, Type 1 A&E. All of England, 2023/24.
tar_read(ae_times_depart_plot)
```

##### Table

```{r}
#| label: tbl-type1-waiting-times
#| tbl-cap: Waiting time for assessment, treatment, conclusion and to leave department, Type 1 A&E. All of England, 2023/24.
tar_read(ae_times_table) |>
  create_dt()
```
:::

#### Mental health frequent attendees

```{r freq_fly_figs, include = FALSE}
avg_all_attends <- tar_read(ae_freq) |> summarise(total = sum(all_pats)) |> pull(total)/5

avg_mh_attend <- tar_read(ae_freq) |> summarise(total = sum(mh_pats)) |> pull(total)/5

perc_mh <- round(avg_mh_attend/avg_all_attends*100,1)

avg_ff_attend <- tar_read(ae_freq) |> summarise(total = sum(mh_freqfly)) |> pull(total)/5

perc_freq <- round(avg_ff_attend/avg_mh_attend*100,1)

```

Frequent attendees can be defined as those attending 5 or more times in a 12 month period with a primary complaint/diagnosis of mental health disorders.

Over the past 5 years in England, on average, there are `r comma(round(avg_mh_attend))` patients presenting with a mental health problem. Of these around `r comma(round(avg_ff_attend))` ( `r perc_freq`% ) are those who attend frequently. This varies by ICB area.

::: panel-tabset
##### Overview

```{r ae_freq_icb_plot}
#| label: fig-type1-freq
#| fig-cap: Percentage of with 5+ attendances in year, Type 1 A&E. All of England, 2023/24.

tar_read(ae_freq_boxplot)

```

##### Summary

```{r}
#| label: tbl-type1-freq-summary
#| tbl-cap: Summary of percentage of with 5+ attendances over time.
tar_read(ae_freq_data) |>
  ungroup() |>
  rename(icb_code = icb23cd,
         value = perc_freq) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-type1-freq
#| tbl-cap: Percentage of Mental Health Type 1 patients attending 5+ times a year by ICB.
tar_read(ae_freq_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 2.5, 0.35))
```
:::

### Detailed breakdowns {#sec-type1-breakdowns}

#### Percentage of all Type 1 attendances with Mental Health as primary reason/diagnosis {#sec-type1-mh-attends}

@fig-type1-perc-mh suggests the median percentage of Type 1 attendances with Mental Health as the primary reason/diagnosis is comparable with pre-pandemic levels. Breakdowns by ICB are in @tbl-type1-perc and breakdowns by population demographics below.

::: panel-tabset
##### Overview

```{r}
#| label: fig-type1-perc-mh
#| fig-cap: Percentage of Mental Health Type 1 attendances over time.
tar_read(type1_mh_attends_boxplot)
```

##### Summary

```{r}
#| label: tbl-type1-perc-summary
#| tbl-cap: Summary of percentage of Mental Health Type 1 attendances over time.
tar_read(type1_perc_mh_attends) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-type1-perc
#| tbl-cap: Percentage of Mental Health Type 1 attendances by ICB.
tar_read(type1_mh_attends_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 2.5, 0.35))
```
:::

#### Percentage of MH Type 1 attendances known to MH services {#sec-type1-mh-known}

The percentages of Mental Health Type 1 attendances where the patient is already known to MH services varies over time (@fig-type1-mh-known-boxplot) and by ICB (@tbl-type1-mh-known).

::: panel-tabset
##### Overview

```{r}
#| label: fig-type1-mh-known-boxplot
#| fig-cap: Percentage of Mental Health Type 1 attendances known to MH services over time.
tar_read(type1_mh_known_boxplot)
```

##### Summary

```{r}
#| label: tbl-type1-mh-known-summary
#| tbl-cap: Summary of percentage of Mental Health Type 1 known to MH services over time.
tar_read(type1_perc_mh_known) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-type1-mh-known
#| tbl-cap: Percentage of Mental Health Type 1 attendances known to MH services by ICB.
tar_read(type1_mh_known_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(60, 90, 4)) 
```
:::

#### Utilisation rates, population sub-groups {#sec-type1-utilisation}

Breakdowns for the number of Type 1 attendances with MH as a primary reason/diagnosis are below.

##### Geographical variation

::: panel-tabset
##### Map

```{r icb_map_allmh}
#| label: fig-type1-map
#| fig-cap: Attendance rate by ICB per 100,000 population, Type 1 A&E. All of England, 2023/24.
tar_read(icb_ed_map_2324)
```

##### Table

```{r}
tar_read(icb_rates_ed) |> 
  filter(der_financial_year == '2023/24') |>
  dplyr::left_join(tar_read(icb_codes_names), "icb_code") |>
  dplyr::select(icb_name, der_financial_year, attends, pop, value, lowercl) |>
    dplyr::select(1:6) |>
    dplyr::arrange(der_financial_year) |>
    dplyr::mutate(across(where(is.numeric), ~janitor::round_half_up(., 2))) |>
  create_dt()
```
:::

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-type1-line
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances. All of England.
tar_read(avg_type1_mh_attends_rate_plot)
```

###### Table

```{r }
tar_read(avg_type1_mh_attends_rate) |>
    dplyr::select(1:6) |>
    dplyr::arrange(der_financial_year) |>
    dplyr::mutate(across(where(is.numeric), ~janitor::round_half_up(., 2))) |>
  create_dt()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r type1_attends_gender_plot}
#| label: fig-type1_attends-gender
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by gender.
tar_read(gender_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_gender_table}
tar_read(gender_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r type1_attends_age_plot}
#| label: fig-type1_attends-age
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by age group.
tar_read(age_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_age_table}
tar_read(age_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r type1_attends_ethnic_plot}
#| label: fig-type1_attends-ethnic
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by ethnic category.
tar_read(ethnic_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_ethnic_table}
tar_read(ethnic_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart1

```{r type1_attends_imd_plot}
#| label: fig-type1_attends-imd
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by Index of Multiple Deprivation decile.
tar_read(imd_plot)$type1_mh_attends
```

###### Chart2

```{r type1_attends_imd_plot2}
#| label: fig-type1_attends-imd2
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by Index of Multiple Deprivation decile and year.
tar_read(ae_attends_imd)
```

###### Table

```{r type1_attends_imd_table}
tar_read(imd_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```
:::

##### Rural-urban variation

::: panel-tabset
###### Chart

```{r type1_attends_rural_plot}
#| label: fig-type1_attends-rural
#| fig-cap: Rates per 100,000 population of the number of Mental Health Type 1 attendances by rural and urban populations.
tar_read(rural_plot)$type1_mh_attends
```

###### Table

```{r type1_attends_rural_table}
tar_read(rural_breakdowns)$type1_mh_attends |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```
:::
:::

## Detailed analysis - Walk-in and Urgent Care attendance {#sec-type34-overview}

```{r uec_summary_bits, include = FALSE}

uec_all <- tar_read(uec_summary) |> filter(der_financial_year == '2023/24') |> pull(all)/1000000

uec_all_cost <- tar_read(uec_summary) |> filter(der_financial_year == '2023/24') |> pull(all_cost)/1000000000

uec_mh <- tar_read(uec_summary) |> filter(der_financial_year == '2023/24') |> pull(mh)/1000

uec_mh_cost <- tar_read(uec_summary) |> filter(der_financial_year == '2023/24') |> pull(mh_cost)/1000000

```

These attendances are those specified within the ECDS as 'Type 3' urgent treatment centre and 'Type 4' walk-in centre attendances and should be typically self-conveyed patients requiring urgent assessment and treatment to prevent deterioration.

The average urgent care attendance in England costs £107 <https://www.england.nhs.uk/costing-in-the-nhs/national-cost-collection/>.

During 2023/24 there were `r round(uec_all,2)` million Type 3/4 attendances costing £ `r round(uec_all_cost,2)` billion. Of those attendances c`r round(uec_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(uec_mh_cost,2)` million.

```{r uec_top_ten_causes}
#| label: tbl-uec-snomed
#| tbl-cap: Top 10 SNOMED codes in UCC and WIC presentation. All of England, 2023/24.

gt_causes <- tar_read(ae_diag) |> 
  filter(der_financial_year == '2023/24', ec_department_type %in% c(3,4)) |> 
  summarise(attends = sum(attends), .by = term) |>
  rename(SNOMED_term = term) |> 
  arrange(desc(attends)) |> 
  slice_head(n = 10) 

create_dt(gt_causes) 
```

### Access and patient pathway {#sec-type34-pathway}

#### Arrival mode to ED

Data is captured in ECDS for the mode of transport used to get the patient to the department categorised to ambulance conveyance, self-conveyance and by public transport.

::: panel-tabset
##### All modes

```{r uec_arrive_mode}
#| label: fig-uec_attends-mode
#| fig-cap: Percentage of attendances by arrival mode, Type 3/4 departments in England, 2023/24.
tar_read(uec_trans_barplot)
```

##### Trend

```{r uec_arrive_mode_trend}
#| label: fig-uec_attends-mode-trend
#| fig-cap: Percentage of attendances by arrival mode, Type 3/4 departments in England, 2019/20 to 2023/24.
tar_read(uec_trans_trends)
```

##### Table

```{r uec_arrive_mode_table}
#| label: tbl-uec-arrival
#| tbl-cap: Number and percentage of attendances by arrival mode and financial year.
tar_read(uec_transp_trends) |>
  create_dt()
```
:::

#### Time of arrival

::: panel-tabset
##### Chart

```{r uec_arrive_time_plot}
#| label: fig-uec-toa
#| fig-cap: Percentage of attendances by hour of arrival, all of England, 2019/20 to 2023/24.
tar_read(uec_toa_plot)
```

##### Table

```{r uec_arrive_time_table}
#| label: tbl-uec-toa
#| tbl-cap: Number and percentage of attendances by hour of arrival, 2019/20 to 2023/24.
tar_read(uec_toa_summary) |>
  create_dt()
```
:::

#### Leaving the department

SNOMED departure status codes allow us to identify patients that left the department before either assessment or treatment was completed.

::: panel-tabset
##### Chart

```{r uec_left_plot}
#| label: fig-uec_attends-left
#| fig-cap: Percentage of attendances left before assessment or treatment, Type 3/4 departments in England, 2023/24.
tar_read(uec_left_chart)
```

##### Table

```{r uec_left_summary}
#| label: tbl-uec-leave
#| tbl-cap: Number of attendances and patients leaving the departments before assessment or treatment by financial year.
tar_read(uec_left_summary) |>
  create_dt()
```
:::

### Detailed breakdowns {#sec-type34-breakdown}

#### Percentage of all UEC attendances with MH as primary reason/diagnosis {#sec-uec-mh-attends}

@fig-uec-perc-mh suggests greater percentage of Mental Health UEC attendances in 2023/24. This varies by ICB (@tbl-uec-perc) where some ICBs have consistently low or high percentages over time.

::: panel-tabset
##### Overview

```{r}
#| label: fig-uec-perc-mh
#| fig-cap: Percentage of Mental Health UEC attendances over time.
tar_read(uec_mh_attends_boxplot)
```

##### Summary

```{r}
#| label: tbl-uec-perc-summary
#| tbl-cap: Summary of percentage of Mental Health UEC attendances over time.
tar_read(uec_perc_mh_attends) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-uec-perc
#| tbl-cap: Percentage of Mental Health UEC attendances by ICB.
tar_read(uec_mh_attends_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(0, 1.8, 0.25))
```
:::

#### Percentage of MH UEC attendances known to MH services {#sec-uec-mh-known}

@fig-uec-mh-known-boxplot suggests lower percentage of Mental Health UEC attends were known in 2023/24. This again varies by ICB (@tbl-uec-mh-known).

::: panel-tabset
##### Overview

```{r}
#| label: fig-uec-mh-known-boxplot
#| fig-cap: Percentage of Mental Health UEC attendances known to MH services over time.
tar_read(uec_mh_known_boxplot)
```

##### Summary

```{r}
#| label: tbl-type1-mh_known-summary
#| tbl-cap: Summary of percentage of Mental Health UEC attendances known to MH services over time.
tar_read(uec_perc_mh_known) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-uec-mh-known
#| tbl-cap: Percentage of Mental Health UEC attendances known to MH services by ICB.
tar_read(uec_mh_known_table) |>
  color_gradient(c("2019/20", "2020/21", "2021/22", "2022/23", "2023/24"), 
                 seq(30, 90, 8)) 
```
:::

#### Utilisation rates, population sub-groups {#sec-type34-populations}

Breakdowns for the number of attendances to Urgent Care or Walk-in with MH as primary reason/diagnosis are below.

##### Geographical variation

::: panel-tabset
##### Map

```{r icb_map_allmh_uec}
#| label: fig-uec_attends-map
#| fig-cap: Attendances rate per 100,000 population, Type 3/4 departments by ICB, 2023/24.
tar_read(icb_uec_map_2324)
```

##### Table

```{r}
tar_read(icb_rates_uec) |> 
  filter(der_financial_year == '2023/24') |>
  dplyr::left_join(tar_read(icb_codes_names), "icb_code") |>
  dplyr::select(icb_name, der_financial_year, attends, pop, value, lowercl) |>
    dplyr::select(1:6) |>
    dplyr::arrange(der_financial_year) |>
    dplyr::mutate(across(where(is.numeric), ~janitor::round_half_up(., 2))) |>
  create_dt()
```
:::

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-uec-line
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances. All of England.
tar_read(avg_uec_mh_attends_rate_plot)
```

###### Table

```{r }
tar_read(avg_uec_mh_attends_rate) |>
    dplyr::select(1:6) |>
    dplyr::arrange(der_financial_year) |>
    dplyr::mutate(across(where(is.numeric), ~janitor::round_half_up(., 2))) |>
  create_dt()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r uec_attends_gender_plot}
#| label: fig-uec-attends-gender
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by gender.
tar_read(gender_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_gender_table}
tar_read(gender_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r uec_attends_age_plot}
#| label: fig-uec-attends-age
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by age group.
tar_read(age_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_age_table}
tar_read(age_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```
:::

##### Ethnic group variation

::: panel-tabset
###### Chart

```{r uec_attends_ethnic_plot}
#| label: fig-uec-attends-ethnic
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by ethnic category.
tar_read(ethnic_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_ethnic_table}
tar_read(ethnic_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("ethnic_category") |>
  create_dt()
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart1

```{r}
#| label: fig-uec-attends-imd_1
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by Index of Multiple Deprivation decile.
tar_read(imd_plot)$uec_mh_attends
```

###### Chart2

```{r uec_attends_imd_plot2}
#| label: fig-uec-attends-imd_2
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by Index of Multiple Deprivation decile and year.
tar_read(uec_attends_imd)
```

###### Table

```{r uec_attends_imd_table}
tar_read(imd_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```
:::

##### Rural-urban variation

::: panel-tabset
###### Chart

```{r uec_attends_rural_plot}
#| label: fig-uec-attends-urban
#| fig-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by rural or urban classification.
tar_read(rural_plot)$uec_mh_attends
```

###### Table

```{r uec_attends_rural_table}
#| label: tbl-uec-attends-rural
#| tbl-cap: Rates per 100,000 population of the number of Mental Health UEC attendances by rural and urban populations.
tar_read(rural_breakdowns)$uec_mh_attends |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```
:::
:::

## Detailed analysis - NHS 111 calls {#sec-nhs111-overview}

The following are a summary of NHS 111 telephone calls data where a caller is connected (i.e. didn't hang up before connection). This includes calls directed via the NHS 111 online service, though data quality does not allow for accurate breakdown of routes into the telephone service. No separate/standalone NHS 111 online data is available at this time though may be useful if available at a future date.

At the time of this analysis, only data up to 2021/22 was accessible, more may become available in the future.

```{r 111_summary_bits, include = FALSE}

nhs111_all <- tar_read(nhs111_summary) |> filter(der_financial_year == '2021/22') |> pull(all)/1000000

nhs111_all_cost <- tar_read(nhs111_summary) |> filter(der_financial_year == '2021/22') |> pull(all_cost)/1000000000

nhs111_mh <- tar_read(nhs111_summary) |> filter(der_financial_year == '2021/22') |> pull(mh)/1000

nhs111_mh_cost <- tar_read(nhs111_summary) |> filter(der_financial_year == '2021/22') |> pull(mh_cost)/1000000

```

Callers to the NHS 111 service would be typically those unable to reach primary care services or to find out information and the appropriate treatment setting for a non-urgent health complaint.

The average cost of a call to NHS 111 is around £12 with those requiring clinical call-back an extra £15 and hear-treat or see-treat ambulance (999 referrals) costing on average another £203 <https://www.ncbi.nlm.nih.gov/books/NBK575169/>.

During 2021/22 there were `r round(nhs111_all,2)` million NHS 111 calls costing £ `r round(nhs111_all_cost,2)` billion. Of those calls c`r round(nhs111_mh,0)`k had a primary complaint/diagnosis of mental health disorders costing around £ `r round(nhs111_mh_cost,2)` million.

```{r 111_top_ten_symptoms}
#| label: tbl-111-symptoms
#| tbl-cap: Top 10 symptom codes in NHS111 calls. All of England, 2021/22.

gt_causes <- tar_read(nhs111_sympt_summary) |> 
  filter(der_financial_year == '2020/21') |>
  ungroup() |> 
  mutate(calls = all) |>
  select(sg_sd_description, calls) |> 
  arrange(desc(calls)) |> 
  slice_head(n = 10) 

create_dt(gt_causes)
```

### Access and patient pathway {#sec-nhs111-pathway}

#### Time of call

There is very little difference in the times of day that patients with mental health issues call NHS 111 compared to those with physical complaints.

::: panel-tabset
##### Chart

```{r 111_time_of_day_calls}
#| label: fig-nhs111-toc
#| fig-cap: Call times by Hour, NHS 111 calls, All of England, 2021/22.
tar_read(nhs111_perc_mh_calls_toa)
```

##### Table

```{r nhs111_disposition_table}
#| label: tbl-nhs111-toc
#| tbl-cap: Call times by Hour, NHS 111 calls, All of England, 2021/22.
tar_read(nhs111_perc_toa) |>
   create_dt()
```
:::

#### Call disposition

::: panel-tabset
##### All modes

```{r nhs111_disposition}
#| label: fig-nhs111-disposal
#| fig-cap: Disposal 'destination', NHS 111 calls, All of England, 2021/22.
tar_read(nhs111_disposition_bar)
```

##### Trend

```{r nhs111_disposition_trend}
#| label: fig-nhs111-disposal-trend
#| fig-cap: Disposal 'destination' by year, NHS 111 calls, All of England, 2017/18 to 2021/22.
tar_read(nhs111_disposition_trends_chart)
```

##### Table

```{r nhs111_disposition_table}
#| label: tbl-nhs111-disposition
#| tbl-cap: Number and percentage of calls by disposal destination and financial year.
tar_read(nhs111_disposition_trends) |>
   create_dt()
```
:::

#### Mental Health frequent callers

```{r nhs111_freq_fly_figs, include = FALSE}
avg_all_attends_nhs111 <- tar_read(nhs111_freq) |> summarise(total = sum(all_pats)) |> pull(total)/5

avg_mh_attend_nhs111 <- tar_read(nhs111_freq) |> summarise(total = sum(mh_pats)) |> pull(total)/5

perc_mh_nhs111 <- round(avg_mh_attend_nhs111/avg_all_attends_nhs111*100,1)

avg_ff_attend_nhs111 <- tar_read(nhs111_freq) |> summarise(total = sum(mh_freqfly)) |> pull(total)/5

perc_freq_nhs111 <- round(avg_ff_attend_nhs111/avg_mh_attend_nhs111*100,1)

```

Frequent callers can be defined as those calling 5 or more times in a 12 month period with a primary complaint/diagnosis of mental health disorders.

Over the past 5 years in England, on average, there are `r comma(round(avg_mh_attend_nhs111))` patients calling NHS 111 with a mental health problem. Of these around `r comma(round(avg_ff_attend_nhs111))` ( `r perc_freq_nhs111`% ) are those who call NHS 111 frequently. This varies by ICB area.

::: panel-tabset
##### Overview

```{r nhs111_freq_icb_plot}
#| label: fig-nhs111-freq
#| fig-cap: Percentage of NHS 111 callers with 5+ contacts in a year, All of England, 2021/22.

tar_read(nhs111_freq_boxplot)

```

##### Summary

```{r}
#| label: tbl-nhs111-freq-summary
#| tbl-cap: Summary of percentage of with 5+ attendances over time.
tar_read(nhs111_freq_data) |>
  ungroup() |>
  rename(icb_code = icb23cd,
         value = perc_freq) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-nhs111-freq
#| tbl-cap: Percentage of NHS 111 callers with 5+ contacts in a year by ICB.
tar_read(nhs111_freq_table) |>
  color_gradient(c("2017/18", "2018/19", "2019/20", "2020/21", "2021/22"), 
                 seq(1, 8, 1))
```
:::

### Detailed breakdowns {#sec-nhs111-breakdown}

#### Percentage of all calls being MH-related

The percentage of NHS 111 calls being MH-related has increased over time (fig-nhs111-perc-mh). Breakdowns by ICB are in @tbl-nhs111-perc and breakdowns by population demographics can be seen below.

::: panel-tabset
##### Overview

```{r}
#| label: fig-nhs111-perc-mh
#| fig-cap: Percentage of Mental Health calls over time.
tar_read(nhs111_mh_calls_boxplot)
```

##### Summary

```{r}
#| label: tbl-nhs111-perc-mh
#| tbl-cap: Summary of percentage of Mental Health calls over time.
tar_read(nhs111_perc_mh_calls) |>
  get_quantiles_and_mean()
```

##### ICB Variation

```{r}
#| label: tbl-nhs111-perc
#| tbl-cap: Percentage of Mental Health Type 1 calls by ICB.
tar_read(nhs111_mh_calls_table) |>
  color_gradient(c("2017/18", "2018/19", "2019/20", "2020/21", "2021/22"), 
                 seq(0, 1.8, 0.25))
```
:::

#### Utilisation rates, population sub-groups {#sec-nhs111-populations}

##### Geographical variation

::: panel-tabset
##### Map

```{r}
#| label: fig-nhs111-icb-map
#| fig-cap: Mental Health calls rate per 100,000 to NHS 111 by ICB of residence, 21/22
tar_read(icb_111_map_2122)
```

##### Table

```{r}
tar_read(nhs111_mh_calls_rate) |> 
  filter(der_financial_year == '2020/21') |>
  dplyr::left_join(tar_read(icb_codes_names), "icb_code") |>
  dplyr::select(icb_name, der_financial_year, attends, pop, value, lowercl) |>
  dplyr::arrange(der_financial_year) |>
  dplyr::mutate(across(where(is.numeric), ~janitor::round_half_up(., 2))) |>
  create_dt()
```
:::

::: panel-tabset
##### Overview

::: panel-tabset
###### Chart

```{r}
#| label: fig-nhs111-line
#| fig-cap: Rates per 100,000 population of the number of Mental Health related calls to NHS111. All of England.
tar_read(avg_nhs111_mh_calls_rate_plot)
```

###### Table

```{r }
tar_read(avg_nhs111_mh_calls_rate) |>
  dplyr::rename(calls = attends) |>
  dplyr::select(1:6) |>
  dplyr::arrange(der_financial_year) |>
  dplyr::mutate(across(where(is.numeric), ~janitor::round_half_up(., 2))) |>
  create_dt()
```
:::

##### Gender variation

::: panel-tabset
###### Chart

```{r nhs111_calls_gender_plot}
#| label: fig-nhs111-calls-gender
#| fig-cap: Rates per 100,000 population of the number of Mental Health related calls to NHS111 by gender.
tar_read(gender_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_gender_table}
tar_read(gender_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("gender") |>
  create_dt()
```
:::

##### Age-group variation

::: panel-tabset
###### Chart

```{r nhs111_calls_age_plot}
#| label: fig-nhs111-calls-age
#| fig-cap: Rates per 100,000 population of the number of Mental Health related calls to NHS111 by age group.
tar_read(age_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_age_table}
tar_read(age_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("age_group") |>
  create_dt()
```
:::

##### Socioeconomic variation

::: panel-tabset
###### Chart1

```{r nhs111_calls_imd_plot}
#| label: fig-nhs111-calls-imd
#| fig-cap: Rates per 100,000 population of the number of Mental Health related calls to NHS111 by Index of Multiple Deprivation decile.
tar_read(imd_plot)$nhs111_mh_calls
```

###### Chart2

```{r nhs111_calls_imd_plot2}
#| label: fig-nhs111-calls-imd2
#| fig-cap: Rates per 100,000 population of the number of Mental Health related calls to NHS111 by Index of Multiple Deprivation decile and year.
tar_read(nhs111_calls_imd)
```

###### Table

```{r nhs111_calls_imd_table}
tar_read(imd_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("imd_decile") |>
  create_dt()
```
:::

##### Rural-urban variation

::: panel-tabset
###### Chart

```{r nhs111_calls_rural_plot}
#| label: fig-nhs111-calls-rural
#| fig-cap: Rates per 100,000 population of the number of Mental Health related calls to NHS111 by rural and urban populations.
tar_read(rural_plot)$nhs111_mh_calls
```

###### Table

```{r nhs111_calls_rural_table}
tar_read(rural_breakdowns)$nhs111_mh_calls |>
  dplyr::rename(calls = attends) |>
  get_standard_table_for_breakdowns("rural_urban") |>
  create_dt()
```
:::
:::
